---
title: "CosMx analysis"
author: "F. Ake"
date: "2026-05-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 7,
  engine.path = list(bash = "/usr/bin/bash")
)
```

## 1. Loading Packages

```{r libraries}

# Import libraries
# ================
library(Seurat)
library(ggplot2)
library(dplyr)
library(future)
library(future.apply)
library(data.table)
library(openxlsx)
library(clustree)
library(ggthemes)
library(Polychrome)
library(scales)
library(colorspace)
library(scPearsonPCA)
library(ggrepel)

# custom functions
source("src/custom_functions.R")

# Global variables
# ================
COSMX_INPUTFILES_REPO = "/scratch/users/franzake/VIVEK_LAB/data/flatFiles"
PANEL_FOV = "Hs_6k"
FOV_MAX_PROP_LOSS = .6

# QC variables
# ============
COUNTS_THRESHOLD_MIN = 30
COUNTS_THRESHOLD_MAX = NULL
AREA_THRESHOLD_MIN = NULL
AREA_THRESHOLD_MAX = 80e3
SBR_THRESHOLD_MIN = 0
SBR_THRESHOLD_MAX = 7.5
FOV_INTEGRITY_PERCENT_THRESHOLD = 0.6
SPLIT_RATIO_THRESHOLD_MIN = NULL
SPLIT_RATIO_THRESHOLD_MAX = log2(4)


# Paralelise heavy steps (SCTransform, RCTD, etc.)
# =================================================
plan("multicore", workers = 6)
options(future.globals.maxSize = 60 * 1024^3)  # 20 GB

```

```{r}
test=gbspatial::dataprep_cosmx("/scratch/users/franzake/VIVEK_LAB/data/flatFiles",plot=T)


tma_mat=t(openxlsx::read.xlsx("data/flatFiles/TA649/TA-649 Autoimmune Hepatitis biopsy.xlsx",sheet = 3,rows = seq(2,20,3),cols = 1:11,colNames = F)[,-1])

library(gbspatial)

tmamapping <- gbspatial::assign_fovs_to_cores(#USC5_247 has issues
  fov_input = list.files("/scratch/users/franzake/VIVEK_LAB/data/flatFiles", pattern = "_fov_positions_file\\.csv\\.gz$",recursive = TRUE, full.names = TRUE),
  slidelabels = unique(test$metadata$Run_Tissue_name),
  cell_input = split(test$metadata, test$metadata$Run_Tissue_name),
  tma_map_input = list(tma_mat)
  )
tmamapping$plots$`TA 649`$Whole_Slide_by_Column
tmamapping$mapped_data[tmamapping$mapped_data$FOV=="TA 649_23",c("core_col","core_row","core_str","id")]=tmamapping$mapped_data[tmamapping$mapped_data$FOV=="TA 649_24",c("core_col","core_row","core_str","id")]
tmamapping$plots$`TA 649`$Whole_Slide_by_Column
View(tmamapping$mapped_data)

tma_mat
```

## 2. Input files

Here are the input files that have to be analyzed...

```{r list_files}
# Sys.which("bash")
# Preview the data directory structure
system("/oak/stanford/groups/bhowitt/conda_envs/workspace/bin/eza -T -L 4 .")
```

## 3. Load CosMx data

```{r data_loading}
# Call the function
usc = readCosMx(experiment_repo = COSMX_INPUTFILES_REPO)
usc

#check TMA & FOvs
TmaPlot(usc, col.by = "FOV")
```

## 4. Quality control

####  - nCounts v nFeatureRNA
Assesses transcriptomic complexity and overall cell quality. 
Cells with very low counts and features are typically low-quality cells, empty segmentations, or background signal.
Extremely high counts may indicate merged cells or segmentation artifacts.

####  - nCounts v Area
Evaluates segmentation consistency. Larger cells are expected to contain more transcripts.
Very small cells often correspond to debris or segmentation fragments, while extremely large cells may represent merged cells or segmentation errors.

#### - SBR v nCounts
Measures transcript detection quality relative to imaging background noise. 
Low SBR values indicate poor-quality or noisy detections and are generally filtered out. 
Extremely high outliers may represent imaging artifacts and should be inspected spatially.

#### - SplitRatioToLocal vs Area
Quality metric associated with cell fragmentation or over-segmentation. 
High values may indicate unstable or incorrect cell boundaries. 
Filtering cells with abnormal split ratios improves transcript assignment accuracy and downstream spatial analysis robustness.


```{r qc_controls_and_filtering}


# Check QC metrics on RAW
# =======================

# Compute FOV integrity
usc = computeFOVintegrity(usc, max_prop_loss = FOV_MAX_PROP_LOSS, panel_name = PANEL_FOV)
# Compute_SBRcomp
usc = computeSBR(usc)
#log2 SplitRatioToLocal
usc$SplitRatioToLocal_log2 = log2(usc$SplitRatioToLocal)

# visualization
# =============
# TmaPlot(usc, col.by = "fov_signal_loss", pt.size = 0.01)
# TmaPlot(usc, col.by = "flag_fov_integrity")
# TmaPlot(subset(usc), col.by = "log2SBR")
# TmaPlot(usc, col.by = "SplitRatioToLocal_log2")


# Check QC metrics with threshold with RAW data
# =============================================
plots = plotQCs(
  usc, 
  count_min = COUNTS_THRESHOLD_MIN, 
  count_max = COUNTS_THRESHOLD_MAX, 
  area_min = AREA_THRESHOLD_MIN, 
  area_max = AREA_THRESHOLD_MAX, 
  sbr_min = SBR_THRESHOLD_MIN, 
  sbr_max = SBR_THRESHOLD_MAX, 
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN, 
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX, 
  base_size = 15) %>% 
  patchwork::wrap_plots(., nrow=4)
plots


# Apply QC filtering developed by Giuseppe
# ========================================
usc_qc <- QCfiltering(
  usc, 
  count_min = COUNTS_THRESHOLD_MIN, 
  count_max = COUNTS_THRESHOLD_MAX, 
  area_min = AREA_THRESHOLD_MIN, 
  area_max = AREA_THRESHOLD_MAX, 
  fov_integrity_threshold = FOV_INTEGRITY_PERCENT_THRESHOLD, 
  sbr_min = SBR_THRESHOLD_MIN, 
  sbr_max = SBR_THRESHOLD_MAX, 
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN, 
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX)
 
plots_qcs = plotQCs(
  usc_qc, 
  count_min = COUNTS_THRESHOLD_MIN, 
  count_max = COUNTS_THRESHOLD_MAX, 
  area_min = AREA_THRESHOLD_MIN, 
  area_max = AREA_THRESHOLD_MAX, 
  sbr_min = SBR_THRESHOLD_MIN, 
  sbr_max = SBR_THRESHOLD_MAX, 
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN, 
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX, 
  base_size = 15) %>% 
  patchwork::wrap_plots(., nrow=4)
plots_qcs

#Upset plots
usc_qc@misc$qc_plots$upset


```


####  - Assignation of FOV to adequate cores

```{r assignation_of_fov_to_cores}

# Global variables & be careful about eventual rotation
# =====================================================
tma_mat = parseTMAmap("data/flatFiles/TA649/TA-649 Autoimmune Hepatitis biopsy.xlsx") %>% t(); tma_mat

#Let'0s replace NA by "Control" here
tma_mat[is.na(tma_mat)] <- "control"

# Extract clinical metadata
# ==========================
usc_qc <- assignCoresToCells(usc_qc,
 tma_map       = tma_mat,
 fov_positions = as.data.frame(fread("data/flatFiles/TA649/TA649_fov_positions_file.csv.gz")), core_drift_tolerance = 0.40
)
usc_qc@misc$fov_core_plots$TA649$Whole_Slide_by_Row
usc_qc@misc$fov_core_plots$TA649$Whole_Slide_by_Column

# Manual Reassing FOV to cores
# ============================
usc_qc$core_str = (usc_qc@meta.data %>% dplyr::mutate(core_str = ifelse(fov==182, "C2R3", core_str)))$core_str
usc_qc$sample_id = (usc_qc@meta.data %>% dplyr::mutate(sample_id = ifelse(fov==182, "37170", sample_id)))$sample_id


# Visualization
# =============

#sampleID assignations
sample_cols = c(
     "#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00",
     "#984EA3", "#A65628", "#F781BF", "#FFED6F",
     "#00CED1", "#8B0000", "#006400", "#FF6347",
     "#4B0082", "#2E8B57"
   )
TmaPlot(usc_qc, col.by = "sample_id", label = T, cols = sample_cols, dark = T)

```



## 5. Dimensionality reduction

### Approach 1 : regular Dimensionality reduction - SCT normalization...

```{r dimensionality_reduction_noBatch_correction}

# Let's employ first Seurat approach (1)
# ==================================

# Find Variable Features
# ----------------------
usc_qc = FindVariableFeatures(usc_qc, nfeatures = 2000)

# Normalization using scTransform approach
# ----------------------------------------
usc_qc = SCTransform(
  usc_qc, 
  assay = "RNA", 
  vars.to.regress = c("nCount_RNA", "nFeature_RNA"),
  clip.range = c(-10, 10))

# PCA computation
# ---------------
usc_qc = RunPCA(usc_qc, npcs = 30, assay = "SCT")
ElbowPlot(usc_qc, ndims = 50)
NPCS = 10

# RunUMAP
# -------
usc_qc = RunUMAP(usc_qc, dims = 1:NPCS, reduction = "pca")
UMAP.plot0 = DimPlot(usc_qc, group.by = "sample_id", label = T, reduction = "umap") +
  ggtitle("UMAP: No Batch correction  - SCT normalization") 
UMAP.plot0

```

Clear Batch effect associated to the sample observed...

### Approach 2 : regular Dimensionality reduction - SCT normalization...

```{r DR_with_harmony_batch_correction}

# Perform Harmony with Batch correction on Sample_ID using prior PCA
# ------------------------------------------------------------------
tmp_00 = usc_qc[["pca"]]
harmony_embedding1 = harmony::RunHarmony(
  usc_qc[["pca"]]@cell.embeddings,
  meta_data = usc_qc@meta.data,
  vars_use = "sample_id")
tmp_00@cell.embeddings = harmony_embedding1
usc_qc[["pca_harmony"]] = tmp_00


# RunUMAP
# -------
usc_qc[["umap_sct_harmony"]] = (RunUMAP(usc_qc, reduction = "pca_harmony", dims = 1:NPCS))@reductions$umap
UMAP.plot1 = DimPlot(usc_qc, group.by = "sample_id", label = T, reduction = "umap_sct_harmony", pt.size = 1.5) +
  ggtitle("UMAP: No Batch correction  - SCT normalization")
UMAP.plot1

```


There is this glob effect thet reflect a poor performance of the SCT norm for the
dimensionality reduction...


Let's employ the Pearson approach from Bruker....

```{r scPearson_approach_with_batch_integration}

DefaultAssay(usc_qc) = "RNA"

# total counts per cell (across all genes)
tc <- Matrix::colSums(usc_qc[["RNA"]]$counts)

# Calculate Gene-Frequency by Batch
genefreq_batch = gene_frequency(
  usc_qc[["RNA"]]$counts, 
  obs = data.table(usc_qc@meta.data)[,.(cell_ID, sample_id)], 
  cellid_colname = "cell_ID", 
  batch_variable = "sample_id")

# RunPCA with Batch effect correction
tmp_00 = (sparse_quasipoisson_pca_seurat_batch(
  usc_qc[["RNA"]]$counts[hvgs,],
  totalcounts = tc,
  grate = genefreq_batch[hvgs,],
  obs =data.table(usc_qc@meta.data)[,.(cell_ID, sample_id)],
  batch_variable = "sample_id", 
  cellid_colname = "cell_ID",
  scale.max = 10,
  do.scale = TRUE,
  do.center = TRUE))
usc_qc[["pearson_pca_batch"]] = tmp_00$reduction.data

# RunPCA with Batch effect
tmp_01 = scPearsonPCA::make_umap(tmp_00)
usc_qc[["pearson_umap_batch"]] = tmp_01$ump
usc_qc[["pearson_graph_batch"]] = Seurat::as.Graph(tmp_01$grph)

# UMAP visualization
UMAP.plot2 = DimPlot(usc_qc, reduction = "pearson_umap_batch", group.by = "sample_id", pt.size = 1.5) +
  ggtitle("UMAP visualization - Batch corrected")
UMAP.plot2

```

Is there any lowQC area that can be identified?

```{r more_on_QC}

lapply(c("nCount_RNA", "nFeature_RNA"), function(x) {
  FeaturePlot(usc_qc, feature = x, order = T, raster=F, assay="SCT") + ggthemes::theme_clean()
}) %>% patchwork::wrap_plots(nrow=2)

```


```{r scPearson_approach_with_harmony_batch_on_pearson_pca}

DefaultAssay(usc_qc) = "RNA"

# total counts per cell (across all genes)
tc <- Matrix::colSums(usc_qc[["RNA"]]$counts)

# PearsonPCA
genefreq <- scPearsonPCA::gene_frequency(usc_qc[["RNA"]]$counts)

# Find Var features
usc_qc = Seurat::FindVariableFeatures(usc_qc, nfeatures = 2000)
hvgs <- VariableFeatures(usc_qc)

# PCA
tmp_02 =sparse_quasipoisson_pca_seurat(
  usc_qc[["RNA"]]$counts[hvgs,],
  totalcounts = tc,
  grate = genefreq[hvgs],
  scale.max = 10,
  do.scale = TRUE,
  do.center = TRUE)

# Perform Harmony with Batch correction on Sample_ID
harmony_embedding2 = harmony::RunHarmony(
  tmp_02$reduction.data@cell.embeddings,
  meta_data = usc_qc@meta.data,
  vars_use = "sample_id")
# reassign embedding to those of Harmony
tmp_02$reduction.data@cell.embeddings = harmony_embedding2

# umap
tmp_03 = scPearsonPCA::make_umap(tmp_02)

# assignation of new embeddings to the right slot in Seurat object
usc_qc[["pearson_pca_harmony"]] = tmp_02$reduction.data


# RunUMAP
usc_qc[["pearson_umap_sct_harmony"]] = (RunUMAP(usc_qc, reduction = "pearson_pca_harmony", dims = 1:13))@reductions$umap
UMAP.plot3 = DimPlot(usc_qc, group.by = "sample_id", label = T, reduction = "pearson_umap_sct_harmony", pt.size = 1.5) +
  theme(legend.position = "none") +
  ggtitle("UMAP: Harmony Batch correction  - scPearson normalization")
UMAP.plot3

```


## 6. Clustering & cell-type characterization

```{r ctype_annotation}

# Find Expression TopMarkers in each clusters
Idents(usc_qc) = usc_qc$seurat_clusters
usc.markers = data.table(FindAllMarkers(usc_qc, assay = "SCT", only.pos = T))
usc.markers


# let's check the average expression of the tops markers in the clusters
top5.markers = (usc.markers %>% group_by(cluster) %>% slice_head(n=30)) %>% data.table()

avg.mat = data.frame(AverageExpression(usc_qc[top5.markers$gene,], group.by = "seurat_clusters", layer = "counts")[["SCT"]])
pheatmap::pheatmap(avg.mat, scale = "row", display_numbers = T)


jci_insight_markers = list(
  "Kupffer_cells/Macrophages" = c("C1QC", "C1QB", "C1QA", "MARCO"),
  "Hepatocytes" = c("SLC27A5", "ADH4", "SLC2A2", "ARG1", "CYP4A11"),
  "HSCs" = c("COL1A1", "COL3A1", "DCN", "TAGLN", "SPARC"),
  "Periportal_hepatocytes" = c("SAA1", "CRP", "FBP1", "ASS1", "PCK1"),
  "Epithelial_cells" = c("KRT18", "PTK2", "SIGMAR1", "VPS29", "MED15"),
  "Stressed_midzonal_hepatocytes" = c("HAMP", "FOS", "JUND", "SERPINE1", "CXCL2"),
  "Plasma_cells_1" = c("IGKC", "IGHG1/2", "JCHAIN", "MZB1", "IGHA1"),
  "IFN_stressed_hepatocytes" = c("CXCL9", "CXCL10", "STAT1", "HLA-G", "HLA-F"),
  "Plasma_cells_2" = c("IGLC1/2", "IGLL5", "IGKC", "IGHG1/2", "JCHAIN"),
  "Pericentral_hepatocytes" = c("GLUL", "TDO2", "FKBP5", "MGST1", "HSD11B1"),
  "Cholangiocytes" = c("KRT7", "KRT8", "TACSTD2", "MMP7", "SPP1")
)

# Generate Feature Plots
plots = lapply(names(jci_insight_markers)[1:2], function(x){
  message(x)
  p = FeaturePlot(usc_qc, features = jci_insight_markers[[x]], order=T, raster=F, combine = F)
  p = lapply(p, function(x) x + theme(legend.position = "none"))
  patchwork::wrap_plots(p, nrow=1)
})

patchwork::wrap_plots(plots, nrow=2)

```













