---
title: "CosMx → Celldega Landscape — TA649"
author: "F. Ake"
date: "2026-09-08"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: inline
---

[Celldega](https://github.com/broadinstitute/celldega) is the Broad's spatial-omics viewer: a Python
pre-processing side (`celldega.pre`) that bakes a dataset into a tiled, web-native
**LandscapeFiles / DegaFiles** directory, and a JS/deck.gl widget (`celldega.viz.Landscape`) that
streams those tiles into Jupyter, VS Code or a browser. It renders every cell and every transcript
of TA649 at once, over the morphology image — which is the thing Seurat's `ImageDimPlot()` cannot do.

**Celldega has no CosMx reader.** `celldega.pre` supports Xenium, MERSCOPE, Visium HD and a
half-finished `"custom"` path. The route the Celldega authors themselves take for CosMx
(`notebooks/CosMx_pre-process.ipynb` in their repo) is to *rewrite the CosMx flat files into the
Xenium file layout* and then drive the ordinary Xenium pipeline with `technology="Xenium"`. That is
what this document does. The `"custom"` path is deliberately avoided: it has no transcript
row-group support, expects a GeoParquet the flat files do not provide, and is less exercised.

Everything heavy here is Python, so the chunks below are `{python}` chunks executed through
`reticulate` against a dedicated venv — same pattern as the scCODA section of `main.Rmd`, for the
same reason (celldega's dependency stack, `pyvips`/`polars`/`pyarrow`, does not belong in the group
R conda env). R chunks are used only for configuration and for exporting annotations out of the
Seurat object.

**Where to run it.** §3–§6 are the build: they stream a 2.6 GB tx file, stitch 182 morphology TIFFs
and write tens of thousands of tiles, so per lab policy they belong in a batch job, not on a login
node. §8–§9 are the viewer and need an interactive kernel. The split works cleanly:

```bash
# build once (RStudio-Server on Slurm, or knit the first six sections headless)
sbatch -p bhowitt --qos=high_p --time=12:00:00 --cpus-per-task=8 --mem=128GB \
  --wrap="Rscript -e \"rmarkdown::render('main2.Rmd')\""
# then attach the widget from a notebook against the DegaFiles the job left behind
```

Once `DEGA/` exists, §8 is cheap — reopening the Landscape re-runs nothing.

## 1. Libraries & Configuration
#### =====================================================================================================

```{r setup, include=FALSE, cache=F, message=FALSE, warning=FALSE}

# 1. Global variables
COSMX_INPUTFILES_REPO = "~/Projects/CosMx_Spatial_transcriptomics/"
SLIDE                 = "TA649"
SEURAT_SAVEPOINT      = paste0(COSMX_INPUTFILES_REPO, "usc2_savepoint.RDS")

# 2. Global chunk options. cache=F throughout: every expensive step below writes a file and
#    skips itself when that file already exists, which is a more honest checkpoint than knitr's
#    cache for multi-GB artefacts that must survive a session restart.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = TRUE,
  cache   = FALSE,
  fig.width = 12,
  fig.height = 8
)

# 3. Load libraries
invisible(
  lapply(
    c("Seurat", "data.table", "reticulate"),
    function(x) { library(x, quietly = T, character.only = T) }
  )
)
```

The venv is built once and reused. `pyvips` needs the libvips **shared library**, which pip cannot
supply — on Sherlock use `ml load system vips` (or `conda install -c conda-forge pyvips libvips`);
on macOS `brew install vips`. Everything else is a wheel.

```{r py_env, cache=F}

# 1. Point reticulate at an isolated celldega venv BEFORE importing anything.
#    Honour $CELLDEGA_PYTHON if the env already exists elsewhere (e.g. a Sherlock micromamba env),
#    otherwise build a local venv. Same escape hatch as $COSMX_PYTHON in src/custom_funct.R.
celldega_python = Sys.getenv("CELLDEGA_PYTHON", unset = NA)

if (!is.na(celldega_python)) {
  use_python(celldega_python, required = TRUE)
} else {
  venv_path = file.path(path.expand(COSMX_INPUTFILES_REPO), ".venv_celldega")
  if (!dir.exists(venv_path)) {
    reticulate::virtualenv_create(venv_path, python = "3.11")
    reticulate::py_install(
      c("celldega", "pyarrow", "polars", "scipy", "scanpy", "anndata",
        "geopandas", "shapely", "tifffile", "imagecodecs", "pyvips"),
      envname = venv_path, pip = TRUE
    )
  }
  use_virtualenv(venv_path, required = TRUE)
}

# 2. Fail loudly rather than half-importing: a missing libvips only surfaces at dzsave time,
#    which is ~40 min into the run.
py_run_string("import celldega, pyvips, polars, pyarrow, tifffile; print('celldega', celldega.__version__)")
```

```{python py_config}

# 1. Paths. FLAT / DECODED are the AtoMx and instrument exports; XEN is the Xenium-shaped
#    intermediate we synthesise; DEGA is what the Landscape widget actually reads.
from pathlib import Path
import os
import celldega as dega

REPO    = Path(os.path.expanduser(r.COSMX_INPUTFILES_REPO))
SLIDE   = r.SLIDE
FLAT    = REPO / "data" / f"{SLIDE}_export" / "flatFiles" / SLIDE
DECODED = REPO / "data" / f"{SLIDE}_export" / "DecodedFiles" / SLIDE / "20260110_014400_S1"
MORPH   = DECODED / "CellStatsDir" / "Morphology2D"

XEN  = REPO / "data" / "celldega" / f"{SLIDE}_xenium"   # Xenium-format intermediates
DEGA = REPO / "data" / "celldega" / f"{SLIDE}_dega"     # DegaFiles consumed by the widget
for p in (XEN, DEGA):
    p.mkdir(parents=True, exist_ok=True)

# 2. Acquisition constants (RunSummary/*_ExptConfig.txt, and the run's fixed tile size).
FOV_PX     = 4256              # ImRows / ImCols
PX_UM      = 0.120280945       # ImPixel_nm / 1000

# 3. Landscape knobs.
#    DOWNSAMPLE sets the resolution of the stitched morphology mosaic AND the coordinate scale
#    written into the affine, so cells/transcripts/image stay locked together by construction.
#    The full-resolution mosaic is ~94k x 131k px = 12.3 Gpx per channel; 4 gives 0.48 um/px and a
#    771 Mpx mosaic, which is plenty for a whole-TMA overview and fits in a laptop's RAM as uint8.
DOWNSAMPLE  = 4
COORD_SCALE = 1.0 / DOWNSAMPLE
TILE_SIZE   = 250              # fine tile size, in mosaic px, for transcript / boundary tiles
MAX_WORKERS = 8

# 4. Morphology channel order. The TIFFs are 5-plane; the plane order is the ChannelID order
#    B,G,Y,R,U from RunSummary/Morphology_ChannelID_Dictionary.txt.
#    (plane, dega_name, button_label, rgb)
CHANNELS = [
    (4, "dapi",     "DNA",      [0,   0,   255]),
    (0, "panck",    "PanCK",    [0,   255, 0  ]),
    (2, "membrane", "Membrane", [255, 0,   255]),
    (3, "cd45",     "CD45",     [255, 255, 0  ]),
]
IMAGE_INFO = [{"name": n, "button_name": b, "color": c} for (_, n, b, c) in CHANNELS]

print("FLAT   ", FLAT)
print("MORPH  ", MORPH, "-", len(list(MORPH.glob("*.TIF"))), "FOV TIFFs")
print("DEGA   ", DEGA)
```

## 2. The coordinate contract
#### =====================================================================================================

Celldega keeps cells, transcripts and image pixels in one **image-pixel** frame, and gets there by
applying a single 3×3 affine (`micron_to_image_transform.csv`) to the vendor's coordinates. So the
whole visualisation rests on one question: *what affine maps CosMx `*_global_px` onto the row/col of
the stitched morphology mosaic?* Guessing it wrong produces a Landscape where cells float next to
their nuclei, which is easy to miss and impossible to interpret.

For TA649 the relation is exact rather than approximate. Checked against
`TA649_fov_positions_file.csv.gz` over all 182 FOVs and all 126,740 rows of the metadata file, with
**zero** residual (not a fit — an identity):

$$x_{global} = x_{fov} + x_{local} \qquad y_{global} = y_{fov} - y_{local}$$

So `y_local` counts *away* from the FOV's `y_global_px` reference line, and each FOV occupies
`y_global ∈ (y_fov − 4256, y_fov]`. Placing tile row `r = y_local` at mosaic row
`(Y_max − y_fov) + r` therefore lays every FOV down consistently, giving the FOV-independent affine

$$col = (x_{global} - X_{min})\cdot s \qquad row = (Y_{max} - y_{global})\cdot s$$

with `s = 1/DOWNSAMPLE`, i.e. the matrix `[[s,0,-X_min·s],[0,-s,Y_max·s],[0,0,1]]`.

The one bit this does **not** pin down is whether raw TIFF row `r` equals `y_local` (top-down) or
`4255 − y_local` (bottom-up) — the vendor documents neither, and the Minerva stitcher in the sibling
`LiverCosMX/` workspace assumes bottom-up (it flips each tile, then inverts Y globally, which
composes back to the same net orientation). Chunk 2b settles it directly against DAPI signal instead
of inheriting the assumption.

```{python coord_contract}

# 1. Verify the local -> global relation exactly, over every cell. This is an assertion, not a fit:
#    if AtoMx ever changes the convention the run stops here rather than producing a subtly
#    misregistered Landscape.
import polars as pl
import numpy as np

fov_pos = pl.read_csv(FLAT / f"{SLIDE}_fov_positions_file.csv.gz").select(
    pl.col("FOV").cast(pl.Int32).alias("fov"),
    pl.col("x_global_px").cast(pl.Float64).alias("x_fov"),
    pl.col("y_global_px").cast(pl.Float64).alias("y_fov"),
)

meta = pl.read_csv(
    FLAT / f"{SLIDE}_metadata_file.csv.gz",
    columns=["fov", "cell_id", "CenterX_local_px", "CenterY_local_px",
             "CenterX_global_px", "CenterY_global_px"],
).with_columns(pl.col("fov").cast(pl.Int32)).join(fov_pos, on="fov", how="left")

res_x = (meta["CenterX_global_px"] - (meta["x_fov"] + meta["CenterX_local_px"])).abs().max()
res_y = (meta["CenterY_global_px"] - (meta["y_fov"] - meta["CenterY_local_px"])).abs().max()
print(f"cells {meta.height:,}   max|x residual| = {res_x}   max|y residual| = {res_y}")
assert res_x == 0 and res_y == 0, "local->global convention changed; re-derive the affine"

# 2. Mosaic geometry and the affine that follows from it.
X_MIN = float(fov_pos["x_fov"].min())
Y_MAX = float(fov_pos["y_fov"].max())
MOSAIC_W_FULL = int(fov_pos["x_fov"].max() - X_MIN) + FOV_PX
MOSAIC_H_FULL = int(Y_MAX - fov_pos["y_fov"].min()) + FOV_PX
MOSAIC_W = int(np.ceil(MOSAIC_W_FULL / DOWNSAMPLE))
MOSAIC_H = int(np.ceil(MOSAIC_H_FULL / DOWNSAMPLE))

AFFINE = np.array([
    [ COORD_SCALE, 0.0,          -X_MIN * COORD_SCALE],
    [ 0.0,        -COORD_SCALE,   Y_MAX * COORD_SCALE],
    [ 0.0,         0.0,           1.0                ],
])
print(f"mosaic full {MOSAIC_W_FULL} x {MOSAIC_H_FULL} px -> /{DOWNSAMPLE} = {MOSAIC_W} x {MOSAIC_H}")
print(f"resolution  {PX_UM * DOWNSAMPLE:.3f} um/px")
print(AFFINE)
```

```{python orientation_check}

# 1. Decide TIFF row orientation empirically: bin cell centroids and the DAPI plane onto the same
#    coarse grid per FOV and correlate, once with y_local read as a top-down row and once bottom-up.
#    Whichever wins is the truth for this run.
import tifffile

def _corr(a, b):
    a = a.ravel() - a.mean(); b = b.ravel() - b.mean()
    return float((a * b).sum() / np.sqrt((a * a).sum() * (b * b).sum() + 1e-12))

DAPI_PLANE = CHANNELS[0][0]
B = 64                                       # bin edge, px -> 66 x 66 grid per FOV
probe_fovs = [1, 20, 45, 77, 100, 130, 160]
score = {"topdown": 0.0, "bottomup": 0.0}

for fv in probe_fovs:
    tif = next(MORPH.glob(f"*_F{fv:05d}.TIF"), None)
    if tif is None:
        continue
    dapi = tifffile.imread(tif)[DAPI_PLANE].astype(np.float32)
    n    = FOV_PX // B
    img  = dapi[:n*B, :n*B].reshape(n, B, n, B).mean(axis=(1, 3))

    pts = meta.filter(pl.col("fov") == fv).select("CenterX_local_px", "CenterY_local_px").to_numpy()
    if len(pts) < 50:
        continue
    dens = np.zeros((n, n), np.float32)
    np.add.at(dens,
              (np.clip((pts[:, 1] // B).astype(int), 0, n-1),
               np.clip((pts[:, 0] // B).astype(int), 0, n-1)), 1.0)

    score["topdown"]  += _corr(dens,       img)
    score["bottomup"] += _corr(dens[::-1], img)

FLIP_TILE = score["bottomup"] > score["topdown"]
print(f"summed density-vs-DAPI correlation: top-down {score['topdown']:.2f}  "
      f"bottom-up {score['bottomup']:.2f}")
print(f"-> y_local is {'BOTTOM-UP (flip each tile)' if FLIP_TILE else 'TOP-DOWN (no flip)'}")
assert max(score.values()) > 1.0, "no orientation correlates with DAPI - check the channel map"
```

On the TA649 export this comes out decisively **top-down** (summed r ≈ 3.7 vs ≈ 0.0 across seven
FOVs): the raw TIFF row *is* `y_local`, and no per-tile flip is needed. The `FLIP_TILE` flag is
carried through to the stitcher anyway so a re-export with the other convention still works.

## 3. Flat files → Xenium-format intermediates
#### =====================================================================================================

Four artefacts, written into `XEN/`, are what the Xenium pipeline reads:

| Xenium file | built from | columns |
|---|---|---|
| `cells.csv.gz` | `*_metadata_file.csv.gz` | `cell_id, x_centroid, y_centroid` |
| `cell_boundaries.parquet` | `*-polygons.csv.gz` | `cell_id, vertex_x, vertex_y` |
| `cell_feature_matrix/` | `*_exprMat_file.csv.gz` | 10x MTX triplet, genes × cells |
| `transcripts.parquet` | `*_tx_file.csv.gz` | `cell_id, transcript_id, feature_name, x_location, y_location` |

Coordinates stay in raw `global_px` here; the affine from §2 is applied later by celldega itself.

**Control features.** The Celldega CosMx notebook drops everything matching `Negative*` /
`SystemControl*`. On this panel that also swallows `NegativeAdd`, which `plex-6atoipi5h2.txt`
classes as *Endogenous* and which NanoString's own `nCount_RNA` counts as RNA — the same quirk
documented in `CLAUDE.md` for `readCosMxAnnData()`. Here we follow the vendor: drop
`^Negative[0-9]+$` (20) and `^SystemControl` (323), keep `NegativeAdd`, leaving **6216** genes.

### a. Cell centroids and polygons

```{python xen_cells}

# 1. cells.csv.gz  <- metadata centroids. Xenium's reader wants exactly these three columns.
#    polars' write_csv does NOT compress based on the file extension - it would leave a plain CSV
#    called .gz, which pandas then tries to gunzip and dies on. Write through gzip explicitly.
import gzip

cells_csv = XEN / "cells.csv.gz"
if not cells_csv.exists():
    cells = meta.select(
        pl.col("cell_id"),
        pl.col("CenterX_global_px").alias("x_centroid"),
        pl.col("CenterY_global_px").alias("y_centroid"),
    )
    with gzip.open(cells_csv, "wb") as fh:
        cells.write_csv(fh)
print("cells.csv.gz", cells_csv.stat().st_size // 1024, "KB",
      "(gzip magic:", open(cells_csv, "rb").read(2) == b"\x1f\x8b", ")")

# 2. cell_boundaries.parquet <- polygon vertices, used as-is.
#    The Celldega CosMx notebook re-centres every polygon onto its metadata centroid, because on
#    their WTx colon sample the polygon positions were spread ~2x within each FOV. That is NOT true
#    here: measured over all 126,740 cells the polygon vertex-mean sits -0.1 +/- 11.8 px from the
#    metadata centroid in x and -0.8 +/- 11.8 px in y (~1.4 um, the ordinary vertex-mean-vs-centroid
#    difference), with slope 1.00. Re-centring would be a no-op at best and would discard the
#    vendor's own centroid at worst, so polygons pass through and the property is asserted instead.
bounds_pq = XEN / "cell_boundaries.parquet"
if not bounds_pq.exists():
    poly = pl.read_csv(
        FLAT / f"{SLIDE}-polygons.csv.gz",
        columns=["cell", "x_global_px", "y_global_px"],
        schema_overrides={"x_global_px": pl.Float64, "y_global_px": pl.Float64},
    ).rename({"cell": "cell_id", "x_global_px": "vertex_x", "y_global_px": "vertex_y"})

    # shapely's LinearRing needs >= 4 coordinates; drop degenerate outlines
    keep = poly.group_by("cell_id").len().filter(pl.col("len") >= 4).select("cell_id")
    poly = poly.join(keep, on="cell_id", how="inner")

    chk = (poly.group_by("cell_id")
               .agg(pl.col("vertex_x").mean().alias("px"), pl.col("vertex_y").mean().alias("py"))
               .join(meta.select(pl.col("cell_id"),
                                 pl.col("CenterX_global_px").alias("mx"),
                                 pl.col("CenterY_global_px").alias("my")),
                     on="cell_id", how="inner"))
    dx = (chk["px"] - chk["mx"]).to_numpy(); dy = (chk["py"] - chk["my"]).to_numpy()
    print(f"polygon centroid - metadata centroid: dx {dx.mean():+.1f} +/- {dx.std():.1f} px, "
          f"dy {dy.mean():+.1f} +/- {dy.std():.1f} px  (n={len(dx):,})")
    assert abs(dx.mean()) < 50 and abs(dy.mean()) < 50, \
        "polygons are offset from centroids - re-centre them (see Celldega CosMx notebook)"

    poly.write_parquet(bounds_pq)
    print(f"{poly.height:,} vertices, {chk.height:,} cells -> {bounds_pq.name}")
else:
    print("cell_boundaries.parquet exists, skipping")
```

### b. Expression matrix → 10x MTX

`read_cbg_mtx()` expects the 10x triplet (`matrix.mtx.gz` genes × cells, `barcodes.tsv.gz`,
`features.tsv.gz`). The exprMat CSV is 142 MB gzipped / 6561 columns wide, so it is parsed with
pyarrow batch-by-batch rather than pandas. Cell barcodes are rebuilt as `c_<slide>_<fov>_<cell_ID>`
to match `cell_id` in the metadata.

```{python xen_cbg}

# 1. Stream the exprMat, keeping only vendor-endogenous columns.
import gzip, re, shutil, time
import pyarrow as pa, pyarrow.csv as pacsv
import scipy.sparse as sp
from scipy.io import mmwrite

CBG_DIR = XEN / "cell_feature_matrix"
NEG_RE  = re.compile(r"^Negative[0-9]+$")

if not (CBG_DIR / "matrix.mtx.gz").exists():
    CBG_DIR.mkdir(parents=True, exist_ok=True)
    with gzip.open(FLAT / f"{SLIDE}_exprMat_file.csv.gz", "rt") as f:
        header = f.readline().strip().split(",")
    genes = [g for g in header[2:]
             if not NEG_RE.match(g) and not g.startswith("SystemControl")]
    print(f"kept {len(genes)} endogenous of {len(header)-2} features "
          f"(NegativeAdd kept: {'NegativeAdd' in genes})")

    col_types = {"fov": pa.int32(), "cell_ID": pa.int64()}
    col_types.update({g: pa.int32() for g in genes})
    reader = pacsv.open_csv(
        FLAT / f"{SLIDE}_exprMat_file.csv.gz",
        read_options=pacsv.ReadOptions(block_size=128 << 20),
        convert_options=pacsv.ConvertOptions(include_columns=["fov", "cell_ID"] + genes,
                                             column_types=col_types),
    )

    slide_no = 1                                  # cell_id is c_<slide>_<fov>_<cell_ID>
    barcodes, blocks, t = [], [], time.time()
    for batch in reader:
        fov = batch.column("fov").to_numpy()
        cid = batch.column("cell_ID").to_numpy()
        barcodes.extend(f"c_{slide_no}_{f}_{c}" for f, c in zip(fov, cid))
        mat = np.empty((batch.num_rows, len(genes)), dtype=np.int32)
        for j, g in enumerate(genes):
            mat[:, j] = batch.column(g).to_numpy(zero_copy_only=False)
        blocks.append(sp.csr_matrix(mat))
    cbg_mat = sp.vstack(blocks).tocsr()
    print(f"cbg {cbg_mat.shape} nnz {cbg_mat.nnz:,}  ({time.time()-t:.0f}s)")

    # 2. Write the triplet. mmwrite wants genes x cells.
    with gzip.open(CBG_DIR / "barcodes.tsv.gz", "wt") as f:
        f.write("\n".join(barcodes) + "\n")
    with gzip.open(CBG_DIR / "features.tsv.gz", "wt") as f:
        f.write("\n".join(f"{g}\t{g}\tGene Expression" for g in genes) + "\n")
    mtx = CBG_DIR / "matrix.mtx"
    mmwrite(str(mtx), cbg_mat.T.tocoo(), field="integer")
    with open(mtx, "rb") as fi, gzip.open(str(mtx) + ".gz", "wb") as fo:
        shutil.copyfileobj(fi, fo, length=16 << 20)
    mtx.unlink()
    del blocks, cbg_mat
else:
    print("cell_feature_matrix/ exists, skipping")
```

### c. Transcripts

The tx file is 2.6 GB gzipped and never fits in memory — `CLAUDE.md` is explicit about this. pyarrow
streams it and writes large row groups, which the tiler in §6 then reads back in bounded memory.
Transcripts with `cell = c_<slide>_<fov>_0` are extracellular; they are kept, because seeing
unassigned transcripts is half the point of looking at a Landscape.

```{python xen_trx}

# 1. Stream tx -> Xenium transcripts.parquet, dropping the same control targets as the CBG.
import pyarrow.compute as pc, pyarrow.parquet as pq

trx_pq = XEN / "transcripts.parquet"
if not trx_pq.exists():
    out_schema = pa.schema([("cell_id", pa.string()), ("transcript_id", pa.int64()),
                            ("feature_name", pa.string()),
                            ("x_location", pa.float64()), ("y_location", pa.float64())])
    reader = pacsv.open_csv(
        FLAT / f"{SLIDE}_tx_file.csv.gz",
        read_options=pacsv.ReadOptions(block_size=256 << 20),
        convert_options=pacsv.ConvertOptions(
            include_columns=["cell", "target", "x_global_px", "y_global_px"],
            column_types={"x_global_px": pa.float64(), "y_global_px": pa.float64()}),
    )
    writer, tid, nrows, t = pq.ParquetWriter(trx_pq, out_schema, compression="zstd"), 0, 0, time.time()
    for i, batch in enumerate(reader):
        tgt  = batch.column("target")
        # pyarrow has no regex match_substring for ^Negative[0-9]+$ cheaply; match_substring_regex does
        keep = pc.invert(pc.or_(pc.match_substring_regex(tgt, r"^Negative[0-9]+$"),
                                pc.starts_with(tgt, "SystemControl")))
        b = batch.filter(keep)
        if b.num_rows == 0:
            continue
        writer.write_batch(pa.record_batch(
            [b.column("cell"),
             pa.array(range(tid, tid + b.num_rows), type=pa.int64()),
             b.column("target"), b.column("x_global_px"), b.column("y_global_px")],
            schema=out_schema))
        tid += b.num_rows; nrows += b.num_rows
        if i % 50 == 0:
            print(f"  {nrows:,} rows  {time.time()-t:.0f}s", flush=True)
    writer.close()
    print(f"DONE {nrows:,} transcripts -> {trx_pq.name}  ({time.time()-t:.0f}s)")
else:
    print("transcripts.parquet exists, skipping")
```

## 4. Annotations from the Seurat savepoint
#### =====================================================================================================

Leiden-on-the-spot would throw away the whole `main.Rmd` pipeline. Instead the labels that took the
work — DeSpotX-decontaminated cell types, the TMA core / donor assignment from `parseTMAmap()` —
are exported and handed to Celldega as the cell clusterings. `celltype_despotx` becomes the
*default* colouring (Celldega's Xenium path reads
`analysis/clustering/gene_expression_graphclust/clusters.csv`); the rest are attached in §7 as
additional selectable clusterings.

```{r export_annotations, cache=F}

# 1. Pull only the metadata columns; the object itself is 600 MB and nothing else is needed here.
usc2 = readRDS(SEURAT_SAVEPOINT)
md   = as.data.table(usc2@meta.data, keep.rownames = "cell_id")

# 2. Take whichever annotation columns this savepoint actually carries, in preference order for
#    the default. main.Rmd has produced celltype_despotx and merged_celltypes at different stages.
ANNOT_COLS = intersect(
  c("celltype_despotx", "merged_celltypes", "sample_id", "core_str", "fov", "seurat_clusters"),
  colnames(md)
)
DEFAULT_ANNOT = ANNOT_COLS[1]
message("exporting annotations: ", paste(ANNOT_COLS, collapse = ", "),
        "  |  default = ", DEFAULT_ANNOT)

annot = md[, c("cell_id", ANNOT_COLS), with = FALSE]
annot[, (ANNOT_COLS) := lapply(.SD, as.character), .SDcols = ANNOT_COLS]

annot_dir = file.path(path.expand(COSMX_INPUTFILES_REPO), "data", "celldega")
dir.create(annot_dir, recursive = TRUE, showWarnings = FALSE)
fwrite(annot, file.path(annot_dir, "cell_annotations.csv"))

# 3. Carry over the one palette main.Rmd actually fixes — the per-donor sample colours — so the
#    Landscape agrees with every TmaPlot already in the manuscript. Cell-type colours were never
#    pinned there, so those are left to Celldega's own generator rather than invented here.
base_sample_cols = c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00",
  "#984EA3", "#A65628", "#F781BF", "#FFED6F",
  "#00CED1", "#8B0000", "#006400", "#FF6347",
  "#4B0082", "#2E8B57", "orange"
)
SAMPLE_COLORS = list()
if ("sample_id" %in% ANNOT_COLS) {
  lv = sort(unique(annot$sample_id))
  SAMPLE_COLORS = as.list(setNames(
    grDevices::rgb(t(grDevices::col2rgb(base_sample_cols[seq_along(lv)])), maxColorValue = 255),
    lv
  ))
}

rm(usc2); invisible(gc())
```

```{python annot_to_clusters}

# 1. Materialise the default annotation as the 10x-style clusters.csv the Xenium path expects.
#    Cells present in the flat files but absent from the (QC-filtered) Seurat object still have to
#    appear, or create_cluster_and_meta_cluster() drops them out of the Landscape entirely - they
#    are labelled "filtered" so their absence from the analysis stays visible on screen.
import pandas as pd

ANNOT_CSV     = REPO / "data" / "celldega" / "cell_annotations.csv"
DEFAULT_ANNOT = r.DEFAULT_ANNOT
ANNOT_COLS    = list(r.ANNOT_COLS)

annot = pd.read_csv(ANNOT_CSV, dtype=str).set_index("cell_id")
all_cells = pd.read_csv(cells_csv, usecols=["cell_id"], dtype=str)["cell_id"]

clusters_csv = XEN / "analysis" / "clustering" / "gene_expression_graphclust" / "clusters.csv"
clusters_csv.parent.mkdir(parents=True, exist_ok=True)
lab = annot[DEFAULT_ANNOT].reindex(all_cells).fillna("filtered")
pd.DataFrame({"Barcode": all_cells.values, "Cluster": lab.values}).to_csv(clusters_csv, index=False)

n_filtered = int((lab == "filtered").sum())
print(f"{DEFAULT_ANNOT}: {lab.nunique()} levels over {len(lab):,} cells "
      f"({n_filtered:,} not in the Seurat object -> 'filtered')")
print(lab.value_counts().head(15))
```

## 5. Morphology mosaic → DeepZoom pyramids
#### =====================================================================================================

Celldega wants one DeepZoom pyramid per displayable channel. There is no stitched image in this
export — only 182 per-FOV 5-plane TIFFs — so the mosaic is built here, at `DOWNSAMPLE`, straight
from the §2 placement rule. (The sibling `LiverCosMX/minerva/minerva_make_ometiff.R` also produces a
stitched OME-TIFF, via terra + bftools; it is not reused because it bakes in its own downsample and
an inverted-Y world frame, and re-deriving the mapping onto it is strictly more work than laying the
tiles down here with a rule we have already verified to the pixel.)

Each channel is assembled into a disk-backed `np.memmap` rather than RAM: at `DOWNSAMPLE=4` a
channel is ~771 MB as uint8, and holding one at a time keeps the peak well under a laptop's memory.

```{python image_pyramids}

# 1. Stitch and pyramidise one channel at a time.
import pyvips

PYR = DEGA / "pyramid_images"
PYR.mkdir(parents=True, exist_ok=True)
SCRATCH = Path(os.environ.get("L_SCRATCH", XEN))       # high-IOPS temp on Sherlock

fov_rows = {int(r_["fov"]): (r_["x_fov"], r_["y_fov"]) for r_ in fov_pos.to_dicts()}
tif_of   = {int(re.search(r"_F(\d+)\.TIF$", p.name, re.I).group(1)): p
            for p in MORPH.glob("*.TIF")}
missing  = sorted(set(fov_rows) - set(tif_of))
assert not missing, f"no morphology TIFF for FOV(s) {missing}"

for plane, name, _btn, _rgb in CHANNELS:
    if (PYR / f"{name}.dzi").exists():
        print(f"{name}: pyramid exists, skipping"); continue

    mm_path = SCRATCH / f"_mosaic_{name}.u8"
    mosaic  = np.memmap(mm_path, dtype=np.uint8, mode="w+", shape=(MOSAIC_H, MOSAIC_W))
    t = time.time()

    # 2. Contrast: one global window per channel, from a coarse sample of every 12th FOV, so all
    #    FOVs share a stretch and tile seams do not turn into brightness steps.
    samp = np.concatenate([
        tifffile.imread(tif_of[f])[plane][::16, ::16].ravel()
        for f in sorted(fov_rows)[::12]
    ])
    samp = samp[samp > 0]
    lo, hi = (np.percentile(samp, (1, 99.5)) if samp.size else (0.0, 1.0))
    hi = float(max(hi, lo + 1)); lo = float(lo)
    del samp

    # 3. Lay every FOV down at its mosaic offset. row0 of the (possibly flipped) tile goes to
    #    Y_MAX - y_fov; see the affine in section 2.
    for k, fv in enumerate(sorted(fov_rows)):
        x_fov, y_fov = fov_rows[fv]
        tile = tifffile.imread(tif_of[fv])[plane]
        if FLIP_TILE:
            tile = tile[::-1]
        tile = tile[::DOWNSAMPLE, ::DOWNSAMPLE].astype(np.float32)
        np.clip(tile, lo, hi, out=tile)
        tile = ((tile - lo) * (255.0 / (hi - lo))).astype(np.uint8)

        r0 = int(round((Y_MAX - y_fov) * COORD_SCALE))
        c0 = int(round((x_fov - X_MIN) * COORD_SCALE))
        h, w = tile.shape
        r1, c1 = min(r0 + h, MOSAIC_H), min(c0 + w, MOSAIC_W)
        # overlapping FOVs: keep the brighter pixel rather than letting the last tile win
        np.maximum(mosaic[r0:r1, c0:c1], tile[:r1-r0, :c1-c0], out=mosaic[r0:r1, c0:c1])
        if (k + 1) % 40 == 0:
            print(f"  {name}: {k+1}/{len(fov_rows)} FOVs  {time.time()-t:.0f}s", flush=True)
    mosaic.flush()

    # 4. DeepZoom. pyvips reads the memmap without copying it into the Python heap.
    vi = pyvips.Image.new_from_memory(memoryview(mosaic), MOSAIC_W, MOSAIC_H, 1, "uchar")
    vi.dzsave(str(PYR / name), tile_size=512, overlap=0, suffix=".webp[Q=90]")
    del vi, mosaic
    mm_path.unlink(missing_ok=True)
    print(f"{name}: {MOSAIC_W}x{MOSAIC_H} pyramid done  ({time.time()-t:.0f}s)", flush=True)

print("max zoom level:", dega.pre.get_max_zoom_level(str(PYR / "dapi_files")))
```

## 6. Celldega pre-processing → DegaFiles
#### =====================================================================================================

From here the ordinary Xenium pipeline runs untouched, with `technology="Xenium"`. Order matters:
`make_meta_gene()` writes the gene→int name mapping that the transcript tiler then looks up, and
`save_landscape_parameters()` has to come last because it inspects everything already on disk.

```{python dega_transform_cells}

# 1. The affine from section 2, in celldega's space-separated 3x3 format.
tpath = DEGA / "micron_to_image_transform.csv"
pd.DataFrame(AFFINE).to_csv(tpath, sep=" ", header=False, index=False)

# 2. Cell centroids -> image px. image_scale stays 1: the scale already lives in the affine, and
#    celldega applies image_scale inconsistently between functions, so keeping it out of play means
#    cells, transcripts and boundaries cannot drift apart.
dega.pre.make_meta_cell_image_coord(
  "Xenium",
  str(tpath),
  str(cells_csv),
  str(DEGA / "cell_metadata.parquet"),
  image_scale = 1
)

cm = pd.read_parquet(DEGA / "cell_metadata.parquet")
xy = np.array(cm["geometry"].to_list())
print(f"{len(cm):,} cells   x {xy[:,0].min():.0f}-{xy[:,0].max():.0f}   "
      f"y {xy[:,1].min():.0f}-{xy[:,1].max():.0f}   (mosaic {MOSAIC_W} x {MOSAIC_H})")
assert xy.min() > -TILE_SIZE and xy[:,0].max() < MOSAIC_W + TILE_SIZE \
       and xy[:,1].max() < MOSAIC_H + TILE_SIZE, "cells fall outside the mosaic - affine is wrong"
```

```{python dega_cbg}

# 1. Read the CBG back, cached as h5ad: the MTX is ~126M nonzeros and slow to re-parse from text.
import anndata as ad

h5 = XEN / "cbg.h5ad"
if h5.exists():
    adata = ad.read_h5ad(h5)
else:
    cbg0  = dega.pre.read_cbg_mtx(str(CBG_DIR), technology="Xenium")
    adata = ad.AnnData(
        sp.csr_matrix(cbg0.sparse.to_coo()).astype(np.float32),
        obs = pd.DataFrame(index=pd.Index(np.asarray(cbg0.index,   dtype=str))),
        var = pd.DataFrame(index=pd.Index(np.asarray(cbg0.columns, dtype=str))),
    )
    adata.obs.index.name = None; adata.var.index.name = None
    adata.write_h5ad(h5); del cbg0

cbg = pd.DataFrame.sparse.from_spmatrix(adata.X, index=adata.obs_names, columns=adata.var_names)
assert str(cbg.index[0]).startswith("c_"), "barcodes are not CosMx cell_ids"
print("cbg", cbg.shape)

# 2. Gene metadata, per-cluster signature matrix, per-gene parquets, cluster colours.
dega.pre.make_meta_gene(cbg, str(DEGA / "meta_gene.parquet"))
dega.pre.cluster_gene_expression("Xenium", str(DEGA), cbg, str(XEN))
dega.pre.save_cbg_gene_parquets("Xenium", str(DEGA), cbg, verbose=False)
dega.pre.create_cluster_and_meta_cluster("Xenium", str(DEGA), str(XEN))

del cbg, adata
import gc; gc.collect()
```

`make_trx_tiles()` loads the whole transcript parquet before tiling, which at CosMx transcript
counts (the tx file alone is 2.6 GB *gzipped*) is an OOM on anything but a `bigmem` node. Celldega's
own CosMx notebook hits the same wall and replaces the call with a streaming tiler; the same is done
here, producing the identical
`transcript_tiles/transcripts_tile_<x>_<y>.parquet` layout (`name` = the int gene id from the §6
mapping, `geometry` = `[x, y]` in mosaic px). Nothing is subsampled.

```{python dega_trx_tiles}

# 1. Gene name -> int, as written by make_meta_gene(). Anything the tiler cannot resolve would be
#    silently dropped, so unresolved targets are counted and reported instead.
from celldega.pre.boundary_tile import _get_name_mapping

gene_to_int = _get_name_mapping(str(DEGA), layer="transcript")

out = DEGA / "transcript_tiles"
if not out.exists():
    spill = Path(os.environ.get("L_SCRATCH", XEN)) / "_trx_spill"
    if spill.exists():
        shutil.rmtree(spill)
    spill.mkdir(parents=True)

    n_tiles_x = int(np.ceil(MOSAIC_W / TILE_SIZE)) + 1
    n_tiles_y = int(np.ceil(MOSAIC_H / TILE_SIZE)) + 1
    seen, dropped, kept, t = set(), 0, 0, time.time()

    for bi, batch in enumerate(pq.ParquetFile(str(trx_pq)).iter_batches(
            batch_size=8_000_000, columns=["feature_name", "x_location", "y_location"])):
        df  = pl.from_arrow(batch)
        uni = df["feature_name"].unique().to_list()
        dropped += int(df.filter(~pl.col("feature_name").is_in(list(gene_to_int))).height)
        df = (df.with_columns(
                  pl.col("feature_name")
                    .replace_strict({g: gene_to_int.get(g) for g in uni},
                                    default=None, return_dtype=pl.Int64).alias("name"))
                .drop_nulls("name"))

        # same affine as the cells: x' = (x - X_MIN)*s ,  y' = (Y_MAX - y)*s
        df = df.with_columns(
                 ((pl.col("x_location") - X_MIN) * COORD_SCALE).round(2).alias("gx"),
                 ((Y_MAX - pl.col("y_location")) * COORD_SCALE).round(2).alias("gy"))
        df = df.with_columns(
                 (pl.col("gx") / TILE_SIZE).floor().cast(pl.Int32).clip(0, n_tiles_x - 1).alias("tx"),
                 (pl.col("gy") / TILE_SIZE).floor().cast(pl.Int32).clip(0, n_tiles_y - 1).alias("ty")
             ).with_columns(pl.concat_list(["gx", "gy"]).alias("geometry"))

        for (tx, ty), td in df.partition_by(["tx", "ty"], as_dict=True).items():
            d = spill / f"{tx}_{ty}"; d.mkdir(exist_ok=True)
            td.select(["name", "geometry"]).write_parquet(d / f"part_{bi:04d}.parquet")
            seen.add((tx, ty))
        kept += df.height
        if bi % 5 == 0:
            print(f"  {kept:,} transcripts, {len(seen)} tiles, {time.time()-t:.0f}s", flush=True)

    # 2. Coalesce the per-batch spill into one parquet per tile.
    out.mkdir()
    for tx, ty in sorted(seen):
        parts = sorted((spill / f"{tx}_{ty}").glob("part_*.parquet"))
        (pl.concat([pl.read_parquet(p) for p in parts]).to_pandas()
           .to_parquet(out / f"transcripts_tile_{tx}_{ty}.parquet", index=False))
    shutil.rmtree(spill)
    print(f"{kept:,} transcripts in {len(seen)} tiles; {dropped:,} unresolved targets "
          f"({time.time()-t:.0f}s)")
    # Reported rather than asserted: this fires after ~an hour of tiling, and a handful of
    # unresolved targets is survivable. On TA649 it is 0 - exprMat and tx use the same gene
    # spelling (both hyphenated, e.g. HLA-DRA), so the '-' vs '_' reconciliation the Celldega
    # CosMx notebook needs for its WTx sample is not required here.
    if dropped:
        print(f"  WARNING: {dropped:,} transcripts dropped - their target is absent from the "
              f"expression matrix. Check gene-name spelling before trusting transcript density.")
else:
    print("transcript_tiles/ exists, skipping")
```

```{python dega_cell_tiles}

# 1. Boundary tiles. tile_bounds must be the mosaic, not the data extent, or the tile grid drifts
#    out of step with the transcript grid above.
tile_bounds = {"x_min": 0.0, "y_min": 0.0, "x_max": float(MOSAIC_W), "y_max": float(MOSAIC_H)}

if not (DEGA / "cell_segmentation").exists():
    dega.pre.make_cell_boundary_tiles(
      "Xenium",
      str(bounds_pq),
      str(DEGA / "cell_segmentation"),
      str(cells_csv),
      str(tpath),
      coarse_tile_factor = 10,
      tile_size          = TILE_SIZE,
      tile_bounds        = tile_bounds,
      image_scale        = 1,
      max_workers        = MAX_WORKERS
    )
else:
    print("cell_segmentation/ exists, skipping")
```

```{python dega_params}

# 1. landscape_parameters.json - written last, it reads the state of everything above.
dega.pre.save_landscape_parameters(
  "Xenium",
  str(DEGA),
  image_name     = "dapi_files",
  tile_size      = TILE_SIZE,
  image_info     = IMAGE_INFO,
  image_format   = ".webp",
  use_int_index  = True,
  use_row_groups = False
)

for p in sorted(DEGA.iterdir()):
    n = len(list(p.iterdir())) if p.is_dir() else ""
    print(f"  {p.name:<32} {'dir, ' + str(n) + ' entries' if p.is_dir() else str(p.stat().st_size // 1024) + ' KB'}")
```

## 7. Switchable annotations (cell type, donor, TMA core, FOV)
#### =====================================================================================================

Two mechanisms exist for colouring cells, and they are not interchangeable:

- **File-based.** `cell_clusters/` in the DegaFiles is the default colouring — that is what §4/§6
  already wrote from `celltype_despotx`. `add_clustering_from_adata(segmentation_name=...)` writes
  extra `cell_clusters_<name>/` directories, but selecting one means setting the widget's
  `segmentation` trait, which *also* redirects it to `meta_gene_<name>.parquet` and
  `cell_segmentation_<name>/`. Those do not exist, so that route 404s unless a whole parallel
  segmentation is duplicated. It is the wrong tool for "recolour by donor".
- **AnnData-based.** `Landscape(adata=...)` takes an AnnData used *only* for cell metadata (never
  expression), keyed on `obs_names` against `cell_metadata.parquet`. `cell_attr` lists the `obs`
  columns to expose and `cluster_attr` picks the active one, switchable from the widget UI.

So the annotation AnnData below is the switchable route, and it is also written to disk so the view
can be reopened without re-running §4.

```{python annot_adata}

# 1. An obs-only AnnData: X is a 1-column zero matrix because Landscape never reads it.
obs = annot.reindex(all_cells).fillna("filtered")
obs.index = all_cells.values
obs.index.name = None
for key in ANNOT_COLS:
    obs[key] = obs[key].astype("category")

ad_annot = ad.AnnData(sp.csr_matrix((len(obs), 1), dtype=np.float32), obs=obs)

# 2. Reuse the donor palette from main.Rmd so the Landscape matches the published TmaPlots.
#    Cell-type colours were never pinned there, so Celldega's deterministic HSV generator supplies
#    those rather than this document inventing a second, conflicting scheme.
# reticulate hands back either a scalar or a length-1 list depending on conversion settings
sample_colors = {k: (v[0] if isinstance(v, (list, tuple)) else v)
                 for k, v in dict(r.SAMPLE_COLORS).items()}
if sample_colors and "sample_id" in ad_annot.obs:
    cats = list(ad_annot.obs["sample_id"].cat.categories)
    ad_annot.uns["sample_id_colors"] = [sample_colors.get(c, "#B0B0B0") for c in cats]

ANNOT_H5AD = DEGA / "cell_annotations.h5ad"
ad_annot.write_h5ad(ANNOT_H5AD)
for key in ANNOT_COLS:
    print(f"  {key:<20} {ad_annot.obs[key].nunique():>4} levels")

# 3. Also refresh the on-disk default clustering, so a statically hosted copy of the DegaFiles
#    (section 10) still opens coloured by cell type with no AnnData in hand.
dega.pre.add_clustering_from_adata(ad_annot, str(DEGA), cluster_key=DEFAULT_ANNOT)
```

## 8. Interactive Landscape
#### =====================================================================================================

The widget reads DegaFiles over HTTP, so they have to be served rather than opened from a path.
`get_local_server()` starts a CORS-enabled thread server **rooted at the current working
directory** and returns its port — so `base_url` is that port plus the path to `DEGA` *relative to
cwd*. For a shared or permanent view, sync `DEGA` to a bucket or GitHub repo and point `base_url` at
the raw URL instead (§10); that is how the Celldega gallery datasets are hosted.

This is an anywidget: it renders in Jupyter, VS Code or a browser, but **will not appear in static
knitted HTML** from RStudio. Run this chunk interactively.

```{python landscape_view}

# 1. Serve the DegaFiles. The server is rooted at cwd, hence the relative path.
port     = dega.viz.get_local_server()
rel_dega = os.path.relpath(DEGA, Path.cwd())
base_url = f"http://localhost:{port}/{rel_dega}"
print("serving DegaFiles at", base_url)

# 2. Open the Landscape. scale_bar_microns_per_pixel keeps the on-screen bar honest: the mosaic is
#    DOWNSAMPLE x coarser than the 0.1203 um instrument pixel.
landscape = dega.viz.Landscape(
  technology   = "Xenium",
  base_url     = base_url,
  adata        = ad_annot,
  cell_attr    = ANNOT_COLS,
  cluster_attr = DEFAULT_ANNOT,
  height       = 800,
  scale_bar_microns_per_pixel = PX_UM * DOWNSAMPLE,
)
landscape
```

```{python landscape_zoom, eval=FALSE}

# 1. Jump straight to one TMA core rather than hunting for it. ini_x / ini_y are mosaic px, so any
#    cell's position out of cell_metadata.parquet is a valid target.
if "core_str" in annot.columns:
    core   = annot.index[annot["core_str"] == "C3R3"]
    sel    = cm[cm["name"].isin(core)]
    cx, cy = np.array(sel["geometry"].to_list()).mean(axis=0)
    landscape.ini_x, landscape.ini_y, landscape.ini_z = float(cx), float(cy), 3.0
    print(f"centred on core C3R3 ({len(sel):,} cells) at ({cx:.0f}, {cy:.0f})")
```

## 9. Landscape + Clustergram
#### =====================================================================================================

`cluster_gene_expression()` in §6 wrote `df_sig.parquet`, the mean expression of every gene in every
level of the default annotation. Rendered as a Clustergram beside the Landscape, clicking a gene row
recolours the tissue by that gene — the marker-checking loop from `main.Rmd`'s dot plots, but on
tissue instead of a UMAP.

```{python clustergram}

# 1. Cluster the signature matrix: column-total normalisation then row z-score, so the map shows
#    enrichment relative to the other cell types rather than absolute abundance.
df_sig = pd.read_parquet(DEGA / "df_sig.parquet")
mat = dega.clust.Matrix(df_sig)
mat.norm(axis="col", by="total")
mat.norm(axis="row", by="zscore")
mat.cluster()

cgm = dega.viz.Clustergram(matrix=mat, width=600, height=700)

# 2. spatial_clustergram() is the current API; landscape_clustergram() is a deprecated alias.
dega.viz.spatial_clustergram(landscape, cgm)
```
## 10. Publishing
#### =====================================================================================================

DegaFiles are static — no server-side code — so a plain object store or a GitHub repo is enough to
share the view. Per `CLAUDE.md`, `$SCRATCH` is purged after 90 days without a write, so the
DegaFiles directory is a build artefact, not the archive: Box (`CosMX AIH/TA649/`) or `$OAK` holds
the source data.

```{bash publish, eval=FALSE}
# 1. Many small tiles - sync, do not tar-and-copy.
ml load system rclone
rclone sync data/celldega/TA649_dega box:"CosMX AIH/TA649/celldega_dega" --transfers 32 --checkers 32

# 2. Then point base_url at the served location:
#    landscape = dega.viz.Landscape(technology='Xenium', base_url='https://.../TA649_dega')
```

```{r session_info}
sessionInfo()
reticulate::py_config()
```
