---
title: "CosMx Data Loading — readCosMx() Test — TA649"
author: "F. Ake"
date: "2026-07-02"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: inline
---


## 1. Libraries & Configuration

Load required packages, source the custom helper library, and define input paths.

```{r setup, include=FALSE, cache=T, message=FALSE, warning=FALSE}


# Set global configs
# ==================
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = TRUE,
  cache   = TRUE,
  fig.width = 7,
  fig.height = 7,
  autodep = TRUE,
  cache.path = "/Users/fake/Projects/LiverCosMX/.cache/knitr/",
  cache.extra = file.mtime("/Users/fake/Projects/LiverCosMX/R/custom_functions.R")
)

# - Load custom functions
# =======================
source("/Users/fake/Projects/LiverCosMX/R/custom_functions.R")

# - Load libraries
# ================
invisible(
  lapply(
    c("Seurat", "data.table", "ggplot2", "patchwork", "dplyr", 
      "ggthemes", "Rmagic", "openxlsx", "celda", "scPearsonPCA", 
      "HieraType", "anndata"), 
    function(x) { library(x, quietly = T, character.only = T) } 
  )
)


# - Define Paths
# ==============
COSMX_INPUTFILES_REPO <- "/Users/fake/Projects/LiverCosMX/data/flatFiles"
TX_FILE       <- Sys.glob(file.path(COSMX_INPUTFILES_REPO, "*", "*_tx_file.csv.gz"))
FOV_POSITIONS <- Sys.glob(file.path(COSMX_INPUTFILES_REPO, "*", "*_fov_positions_file.csv.gz"))
TMA_XLSX      <- Sys.glob(file.path(COSMX_INPUTFILES_REPO, "*", "*.xlsx"))
PANEL_FOV     <- "Hs_6k"

```


## 2. Read CosMx flat files into a Seurat object 

```{r read_cosmx, cache=T}

# - reading Seurat Object from AtomX flatfiles...
usc <- readCosMx(experiment_repo = COSMX_INPUTFILES_REPO)

# - Compute QC metrics within seurat object...
usc <- computeFOVintegrity(usc, max_prop_loss = 0.6, panel_name = PANEL_FOV)
usc <- computeSBR(usc)

```

## 3. Compute QC metrics

```{r qc_metrics, cache=T, fig.height=10}

# QC thresholds
# =============
COUNTS_THRESHOLD_MIN      <- 60        
COUNTS_THRESHOLD_MAX      <- 12000     
FEATURE_THRESHOLD_MIN     <- 60      
FEATURE_THRESHOLD_MAX     <- 4000
AREA_THRESHOLD_MAX        <- 80e3      
SBR_THRESHOLD_MIN         <- 0         
SBR_THRESHOLD_MAX         <- 4.5
SPLIT_RATIO_THRESHOLD_MIN <- log2(0.15) 
SPLIT_RATIO_THRESHOLD_MAX <- log2(4)   
EXTRANUCLEAR_THRESHOLD_MAX <- 98

qc <- QCoverview(
  usc,
  count_min       = COUNTS_THRESHOLD_MIN,
  count_max       = COUNTS_THRESHOLD_MAX,
  feature_min     = FEATURE_THRESHOLD_MIN,
  feature_max     = FEATURE_THRESHOLD_MAX,
  area_max        = AREA_THRESHOLD_MAX,
  sbr_min          = SBR_THRESHOLD_MIN,
  sbr_max          = SBR_THRESHOLD_MAX,
  split_ratio_min  = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max  = SPLIT_RATIO_THRESHOLD_MAX,
  base_size = 13
)

# plot QC panels
# ==============
qc = lapply(
  qc, 
  function(x) x + ggthemes::theme_clean(base_size = 5)
) %>% 
  patchwork::wrap_plots(qc, ncol = 2) ; qc

```

## 5. TMA core assignment

The map is transposed to match the slide layout (`assignCoresToCells()` warns if the orientation
still looks off).

```{r tma_assign, fig.height=6, cache=T}


# Parse the construction key; transpose to match the slide's core grid
# ====================================================================
tma_mat <- t(parseTMAmap(TMA_XLSX))
tma_mat[is.na(tma_mat)] <- "control"
tma_mat

usc <- assignCoresToCells(
  usc,
  tma_map              = tma_mat,
  fov_positions        = as.data.frame(data.table::fread(FOV_POSITIONS)),
  core_drift_tolerance = 0.4, 
  min_cells            = 30
)

# Diagnostic maps to confirm the orientation (row / column sweeps)
# ================================================================
setPlotLabelSize(usc@misc$fov_core_plots[[1]]$Whole_Slide_by_Row, size = 1) +
  ggthemes::theme_clean(base_size = 7)

setPlotLabelSize(usc@misc$fov_core_plots[[1]]$Whole_Slide_by_Column, size = 1) +
  ggthemes::theme_clean(base_size = 7)

# Correction
# ==========
sel <- usc$fov == 23
usc$core_str[sel]  <- "C3R3"
usc$sample_id[sel] <- "37171"


```

```{r tma_plot, cache=T, fig.height=10}

# Visualize FOVs on the TMA using sample-specific colour families
# ===============================================================
base_sample_cols <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00",
  "#984EA3", "#A65628", "#F781BF", "#FFED6F",
  "#00CED1", "#8B0000", "#006400", "#FF6347",
  "#4B0082", "#2E8B57", "orange"
)

sample_levels <- sort(unique(as.character(usc$sample_id)))
sample_cols <- setNames(base_sample_cols[seq_along(sample_levels)], sample_levels)

make_sample_shades <- function(base_col, n) {
  if (n == 1) return(base_col)
  grDevices::colorRampPalette(c(
    colorspace::darken(base_col, 0.20),
    base_col,
    colorspace::lighten(base_col, 0.35)
  ))(n)
}

fov_palette_df <- usc@meta.data %>%
  dplyr::distinct(sample_id, fov, FOV) %>%
  dplyr::mutate(sample_id = as.character(sample_id)) %>%
  dplyr::arrange(sample_id, fov) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(
    fov_col = make_sample_shades(sample_cols[dplyr::first(sample_id)], dplyr::n())
  ) %>%
  dplyr::ungroup()

fov_cols <- setNames(fov_palette_df$fov_col, fov_palette_df$FOV)

# Plotting
# ========
TmaPlot(usc, col.by = 'sample_id', pt.size = 0.05, 
        cols = base_sample_cols, label = T, label.size = 1) +
  theme_clean(base_size=7)

```

## 6. QC filtering

Remove from the Seurat object cells that have been flagged following QC threshold defined earlier

```{r qc_filter, warning=F, cache=T}

usc2 <- QCfiltering(
  usc,
  count_min       = COUNTS_THRESHOLD_MIN,
  count_max       = COUNTS_THRESHOLD_MAX,
  feature_min     = FEATURE_THRESHOLD_MIN,  
  feature_max     = FEATURE_THRESHOLD_MAX,
  area_max        = AREA_THRESHOLD_MAX,
  sbr_min         = SBR_THRESHOLD_MIN,
  sbr_max         = SBR_THRESHOLD_MAX,
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX,
  filter          = TRUE,
  upset_text_scale = 1.3
)

# plotting...
usc2@misc$qc_plots$summary
usc2@misc$qc_plots$upset

```


<!-- Let's perform a decontamination of ambientRNA with DecontX -->

<!-- ```{r ambientRNA_removal, cache=T} -->

<!-- # Run DecontX for detection of ambiantRNA contamination... -->
<!-- dcx_usc2 = celda::decontX( -->
<!--   x    = as(GetAssayData(usc2, assay = "RNA", layer = "counts"), "CsparseMatrix"), -->
<!--   seed = 1 -->
<!-- ) -->
<!-- usc2$contamination <- dcx_usc2$contamination -->

<!-- # Exclude cells with more than 60% of contamination detected with decontX from the object... -->
<!-- usc2[['decontx']] = CreateAssay5Object(counts = dcx_usc2$decontXcounts) -->
<!-- usc2 = subset(usc2, contamination < 0.6) -->

<!-- ``` -->


## 7. Dimensionality reduction with integration

### a. No Ambient RNA Decontamination 

Usage of scPearsonPCA tool for Dimensionality reduction analysis...

```{r dimred, cache=T, fig.height=6}

DefaultAssay(usc2) = "RNA"
LABEL_NAME = "no_decont"

# Usage of scPEarsonPCA for DimReduc
# ==================================

usc2 = runScPearsonDimRed(
  usc2, 
  nfeatures = 700,
  batch_vars = c("sample_id", "fov"),
  label = LABEL_NAME
)

#set default DimReduc
DefaultDimReduc(usc2) = paste0(LABEL_NAME, "_umap")

# 7. Plotting...
p1 = DimPlot(
  usc2, 
  group.by = "sample_id",
  label = F,
  label.size = 4,
  raster = T) +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

p2 = DimPlot(
  usc2, 
  group.by = "clusters",
  label = T,
  label.size = 4,
  raster = T) +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

p1
p2

```

### b. Ambient RNA decontamination

Let's use DeSpotX for generation of a h5ad object

#### - b.1 All cells in Liver

Generation of h5ad file...

```{r seurat_to_h5ad_for_despotx, cache=T}

# Wrappers for generating anndata object
# ======================================
python_anndata <- reticulate::import("anndata", convert = FALSE)
python_pandas <- reticulate::import("pandas", convert = FALSE)

# despotx's Python env (anndata 0.10.9) can't read the "nullable-string-array"
# encoding pandas>=3.0/anndata>=0.13 use by default for string dtypes; force
# plain object-dtype strings so the h5ad stays readable there.
python_pandas$set_option("future.infer_string", FALSE)
make_anndata <- function(X, obs, var, obsm = NULL) {
  obs_names <- rownames(X); if (is.null(obs_names)) obs_names <- rownames(obs)
  var_names <- colnames(X); if (is.null(var_names)) var_names <- rownames(var)
  if (is.data.frame(obs)) rownames(obs) <- obs_names
  if (is.data.frame(var)) rownames(var) <- var_names
  X <- anndata:::.check_matrix(X)   # same matrix-conversion helper the package itself uses
  ad <- python_anndata$AnnData(X = X, obs = obs, var = var, obsm = obsm)  # no dtype arg
  anndata::AnnDataR6$new(ad)         # wrap back into the normal R6 object ($write_h5ad(), $X, etc.)
}

# Conversion of Seurat object into a h5ad object / file
# =====================================================
tc = counts_mat <- Matrix::t(usc2[["RNA"]]$counts)

# 2. Cell metadata — the column DeSpotX will condition on via --cell-type-col.
# sample_id/fov/cell_id are carried through so downstream reclustering
# (runScPearsonDimRed's batch correction) can run on despotx's Sherlock side
# without needing to re-derive them from the original Seurat object.
obs_df <- data.frame(
  cluster   = as.character(usc2$clusters),
  cell_id   = as.character(usc2$cell_id),
  sample_id = as.character(usc2$sample_id),
  fov       = as.character(usc2$fov),
  row.names = colnames(usc2)
)

# 3. Gene metadata — var_names only strictly required by DeSpotX.
var_df <- data.frame(row.names = rownames(usc2[["RNA"]]))

# 4. Spatial coordinates — must be EXACTLY 2D (x, y).
coords <- GetTissueCoordinates(usc2, image = "TA649")
coords <- coords[match(colnames(usc2), coords$cell), c("x", "y")]
spatial_mat <- as.matrix(coords)

# 5. Assemble....
usc2.adata <- make_anndata(
  X    = counts_mat,
  obs  = obs_df,
  var  = var_df,
  obsm = list(spatial = spatial_mat)
)

# 6. Force CSR before writing — via the raw Python object, not the R6 $X<- setter
py_ad <- usc2.adata$.get_py_object()
sp <- reticulate::import("scipy.sparse", convert = FALSE)
py_ad$X <- sp$csr_matrix(py_ad$X)

# sanity check before writing
cat("X format:", as.character(py_ad$X$format), "\n")   # should print "csr"
cat("X shape:", paste(reticulate::py_to_r(py_ad$X$shape), collapse=" x "), "\n")

# 7. Write directly from the raw python object (bypasses the R6 wrapper entirely).
py_ad$write_h5ad("usc2.h5ad")


```

Run DespotX on Sherlock...

```{r run_despotx, cache=T, fig.height=6}

# Ran DespotX on Sherloock for better performances...
# ===================================================
# /Users/fake/Library/Python/3.9/bin/despotx \
#   --h5ad usc2.h5ad \
#   --cell-type-col cluster \
#   --spatial-key spatial \
#   --out despotx_output \
#   --device cuda
  
  
  
# Import adjusted h5ad object and convertion into Seurat object
# =============================================================
usc2.adjusted <- read_h5ad("despotx_output/adjusted.h5ad")

# cells x genes (dgRMatrix) -> genes x cells (dgCMatrix), Seurat's expected orientation
X <- as(usc2.adjusted$X, "CsparseMatrix")
rownames(X) <- rownames(usc2.adjusted$obs)
colnames(X) <- rownames(usc2.adjusted$var)
counts <- Matrix::t(X)

# Integrate into Seurat object
usc2[["despotx"]] = Seurat::CreateAssayObject(counts)

DefaultAssay(usc2) = "despotx"

# Usage of scPEarsonPCA for DimReduc
# ==================================

tc = Matrix::colSums(usc2[["despotx"]]$counts)

# create custom column regarding sample_id & fov as it seems that i observed batch effect related to both cases...
usc2$sample_id_fov = paste0(usc2$sample_id, "-", usc2$fov)

# compute gene frequency...
genefreq_batch = scPearsonPCA::gene_frequency(
  usc2[["despotx"]]$counts, 
  obs = data.table(usc2@meta.data)[,.(cell_id, sample_id_fov)], 
  cellid_colname = "cell_id", 
  batch_variable = c("sample_id_fov")
)

usc2 = FindVariableFeatures(usc2, nfeatures = 2000)
hvgs <- VariableFeatures(usc2, assay = "despotx")

# compute PCA & Bacth integration...
pcaobj_batch = sparse_quasipoisson_pca_seurat_batch(
  usc2[["despotx"]]$counts[hvgs,],
  totalcounts = tc,
  grate = genefreq_batch[hvgs,],
  obs =data.table(usc2@meta.data)[,.(cell_id, sample_id_fov)],
  batch_variable = "sample_id_fov",
  cellid_colname = "cell_id",
  scale.max = 10,
  do.scale = TRUE,
  do.center = TRUE,
  ncores = 8
)

# perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2[["pca_despotx"]] <- pcaobj_batch$reduction.data
usc2[["umap_despotx"]] <- umapobj$ump  
usc2[["graph_despotx"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2 <- Seurat::FindClusters(
  usc2, 
  resolution = 1, 
  graph = "graph_despotx",
  algorithm = 4
)
usc2@meta.data$clusters_despotx <- factor(paste0("G", usc2@meta.data$graph_despotx_res.1))


p1.allcells_despot = DimPlot(
  usc2, 
  group.by = "sample_id",
  label = F,
  label.size = 4,
  reduction = "umap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")

p2.allcells_despot = DimPlot(
  usc2, 
  group.by = "clusters_despotx",
  label = T,
  label.size = 4,
  reduction = "umap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")


# Find Markers...
Idents(usc2) = usc2$clusters_despotx

usc2 <- NormalizeData(
  usc2,
  assay = "despotx",
  normalization.method = "RC",
  verbose = T
)

# Find markers using Seurat FindAllMarkers function...
markers.despotx <- FindAllMarkers(
  usc2,
  assay           = "despotx",
  only.pos        = FALSE,
  min.pct         = 0.05,
  logfc.threshold = 0.1,
  test.use        = "wilcox"
)
fwrite(markers.despotx, file="all.markers.despotx.tsv", sep="\t")

# Find markers using Hieratype ...
markers.despotx.htp = HieraType::clusterwise_foldchange_metrics(
  usc2[["despotx"]]$counts[hvgs,],
  totalcounts = Matrix::colSums(usc2[["despotx"]]$counts),
  metadata = usc2@meta.data, 
  cluster_column = "clusters_despotx")
fwrite(markers.despotx.htp, file="all.markers.despotx.htp.tsv", sep="\t")


# Annotation...
# =============
usc2$celltype_despotx <- dplyr::case_when(
  usc2$clusters_despotx == "G1"  ~ "T/NK.cells",
  usc2$clusters_despotx == "G2"  ~ "Macrophages",

  usc2$clusters_despotx == "G3"  ~ "Hep.Zone1",
  usc2$clusters_despotx == "G4"  ~ "Hep.Zone3",
  usc2$clusters_despotx == "G5"  ~ "Hep.Zone1.2",
  usc2$clusters_despotx == "G8"  ~ "Hep.IFNg",

  usc2$clusters_despotx == "G6"  ~ "Fibroblasts.1/Activated",
  usc2$clusters_despotx == "G7"  ~ "Fibroblasts.2/Quiescent",

  usc2$clusters_despotx == "G9"  ~ "Endothelials (LSEC)",
  usc2$clusters_despotx == "G10" ~ "Cholangiocytes",
  usc2$clusters_despotx == "G11" ~ "Proliferating",

  usc2$clusters_despotx == "G12" ~ "Plasma.cells.1",
  usc2$clusters_despotx == "G13" ~ "Plasma.cells.2",

  usc2$clusters_despotx == "G14" ~ "Monocytes.Neutrophils",

  TRUE ~ "Undefined"
)

celltype_colors_despotx <- c(
  "Hep.Zone3"           = "#08306B",
  "Hep.Zone1"           = "#2171B5",
  "Hep.Zone1.2" = "#4292C6",
  "Hep.IFNg"   = "#6BAED6",

  "Macrophages" = "#E6550D",

  "Fibroblasts.1/Activated" = "#6A3D9A",
  "Fibroblasts.2/Quiescent" = "#A27AC3",

  "Endothelials (LSEC)"          = "#00A087",
  "Cholangiocytes"        = "#D8A600",
  "Proliferating"         = "#525252",
  "T/NK.cells"            = "#33A02C",
  "Plasma.cells.1"        = "#C51B7D",
  "Plasma.cells.2"        = "#DE77AE",
  "Monocytes.Neutrophils" = "#CB181D",

  "Undefined" = "#BDBDBD"
)

# Plotting...
# ===========

p3.allcells_despot = DimPlot(
  usc2, 
  group.by = "celltype_despotx",
  label = T,
  label.size = 3,
  repel = T,
  reduction = "umap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")


p1.allcells_despot
p2.allcells_despot
p3.allcells_despot

```

```{r heatmap_of_markers, cache=T, fig.width=13}

# Heatmap
# =======

markers.despotx.htp = HieraType::clusterwise_foldchange_metrics(
  usc2[["despotx"]]$counts[hvgs,],
  totalcounts = Matrix::colSums(usc2[["despotx"]]$counts),
  metadata = usc2@meta.data, 
  cluster_column = "celltype_despotx")

evidence_genes <- list(
  "Hepatocytes.Zone3"           = c("CYP2E1", "GLUL", "ADH6", "SCD", "MGST1"),
  "Hepatocytes.Zone1"           = c("HAMP", "ASS1", "PCK1", "SDS", "HP"),
  "Hepatocytes.LipidMetabolism" = c("MLXIPL", "PPARA", "PLIN5", "ACSM2A", "GPAM"),
  "Hepatocytes.IFNg.Stressed"   = c("CXCL9", "CXCL10", "GBP1", "STAT1", "MT2A"),

  "Macrophages"   = c("MARCO", "CD5L", "VSIG4", "C1QB", "CD163"),

  "Fibroblasts.1" = c("COL1A1", "COL3A1", "ACTA2", "TAGLN", "TIMP1"),
  "Fibroblasts.2" = c("RELN", "RGS5", "CXCL14", "ECM1", "MASP1"),

  "Endothelials"   = c("STAB1", "KDR", "RAMP3", "CD36", "ENG"),
  "Cholangiocytes" = c("KRT7", "KRT19", "CDH1", "TACSTD2", "AQP1"),
  "Proliferating"  = c("TOP2A", "CDK1", "CCNB1", "PCNA", "STMN1"),

  "T/NK.cells" = c("CD27", "GNLY", "IFNG", "CCL5", "CCL3","GZMA","CD8A","CD8B","CD4"),

  "Plasma.cells.1" = c("IGKC", "IGKV4-1", "JCHAIN", "MZB1", "IGHG1/2"),
  "Plasma.cells.2" = c("IGLC1/2", "IGLL5", "IGLL1", "JCHAIN", "MZB1"),

  "Monocytes.Neutrophils" = c("S100A8", "S100A9", "FCN1", "LYZ", "SRGN")
)

hm_rna <- add_ncells(
  marker_heatmap(
    markers.despotx.htp %>% filter(gene %in% unique(unlist(evidence_genes))), fold_change_min = 0.05
  )
) + 
  labs(title = "Evidence Markers for cell types in CosmX Liver data")


# Plotting...
# ===========
hm_rna

```

Let's focus on specific cell types...

#### - b.2 Macrophages

```{r subset_macrophages, cache=T}

# subset Macrophages
usc2.macrophages = subset(usc2, celltype_despotx %in% c("Macrophages"))


#==============================================================================================
#==============================================================================================

# Perform Dimensionality reduction on subsetted Macrophages & Neutrophiles
# ========================================================================

DefaultAssay(usc2.macrophages) = "despotx"

tc = Matrix::colSums(usc2.macrophages[["despotx"]]$counts)

# 1. Create custom column regarding sample_id & fov as it seems that i observed batch effect related to both cases...
usc2.macrophages$sample_id_fov = paste0(usc2.macrophages$sample_id, "-", usc2.macrophages$fov)

# 2. Compute gene frequency...
genefreq_batch = scPearsonPCA::gene_frequency(
  usc2.macrophages[["despotx"]]$counts, 
  obs = data.table(usc2.macrophages@meta.data)[,.(cell_id, sample_id_fov)], 
  cellid_colname = "cell_id", 
  batch_variable = c("sample_id_fov")
)
usc2.macrophages = FindVariableFeatures(usc2.macrophages, nfeatures = 2000)
hvgs <- VariableFeatures(usc2.macrophages, assay = "despotx")

# 3. compute PCA & Bacth integration...
pcaobj_batch = sparse_quasipoisson_pca_seurat_batch(
  usc2.macrophages[["despotx"]]$counts[hvgs,],
  totalcounts = tc,
  grate = genefreq_batch[hvgs,],
  obs =data.table(usc2.macrophages@meta.data)[,.(cell_id, sample_id_fov)],
  batch_variable = "sample_id_fov",
  cellid_colname = "cell_id",
  scale.max = 10,
  do.scale = TRUE,
  do.center = TRUE,
  ncores = 8
)

# perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2.macrophages[["pca_despotx2"]] <- pcaobj_batch$reduction.data
usc2.macrophages[["umap_despotx2"]] <- umapobj$ump  
usc2.macrophages[["graph_despotx2"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2.macrophages <- Seurat::FindClusters(
  usc2.macrophages, 
  resolution = 1.5, 
  graph = "graph_despotx2",
  algorithm = 4
)
usc2.macrophages@meta.data$clusters <- factor(paste0(
  "G",
  usc2.macrophages@meta.data$graph_despotx2_res.1.5
))

p1.macrophages = DimPlot(
  usc2.macrophages, 
  group.by = "celltype_despotx",
  label = T,
  label.size = 4,
  repel = T,
  raster = T,
  reduction = "umap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none") 

p2.macrophages = DimPlot(
  usc2.macrophages, 
  group.by = "clusters",
  label = T,
  label.size = 4,
  repel = T,
  raster = T,
  reduction = "umap_despotx2") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")


#==============================================================================================
#==============================================================================================

# DecontX decontamination assay
# =============================

dcx <- celda::decontX(
  x    = as(GetAssayData(usc2.macrophages, assay = "despotx", layer = "counts"), "CsparseMatrix"),
  z    = as.character(usc2.macrophages$clusters),
  seed = 1
)
usc2.macrophages$contamination <- dcx$contamination
usc2.macrophages[["decontx"]] <- CreateAssay5Object(counts = dcx$decontXcounts)


#==============================================================================================
#==============================================================================================


# Conversion of Seurat object into a h5ad object / file
# =====================================================
tc = counts_mat <- Matrix::t(usc2.macrophages[["RNA"]]$counts)

# 2. Cell metadata — the column DeSpotX will condition on via --cell-type-col.
obs_df <- data.frame(
  cluster = as.character(usc2.macrophages$clusters),
  row.names = colnames(usc2.macrophages)
)

# 3. Gene metadata — var_names only strictly required by DeSpotX.
var_df <- data.frame(row.names = rownames(usc2.macrophages[["RNA"]]))

# 4. Spatial coordinates — must be EXACTLY 2D (x, y).
coords <- GetTissueCoordinates(usc2.macrophages, image = "TA649")
coords <- coords[match(colnames(usc2.macrophages), coords$cell), c("x", "y")]
spatial_mat <- as.matrix(coords)

# 5. Assemble....
usc2.macrophages.adata <- make_anndata(
  X    = counts_mat,
  obs  = obs_df,
  var  = var_df,
  obsm = list(spatial = spatial_mat)
)

# 6. Force CSR before writing — via the raw Python object, not the R6 $X<- setter
py_ad <- usc2.macrophages.adata$.get_py_object()
sp <- reticulate::import("scipy.sparse", convert = FALSE)
py_ad$X <- sp$csr_matrix(py_ad$X)

# sanity check before writing
cat("X format:", as.character(py_ad$X$format), "\n")   # should print "csr"
cat("X shape:", paste(reticulate::py_to_r(py_ad$X$shape), collapse=" x "), "\n")

# 7. Write directly from the raw python object (bypasses the R6 wrapper entirely).
py_ad$write_h5ad("usc2.macrophages.h5ad")


# 8. Import adjusted h5ad object and convertion into Seurat object
# =============================================================
usc2.macrophages.adjusted <- read_h5ad("despotx_output/macrophages/adjusted.h5ad")

# cells x genes (dgRMatrix) -> genes x cells (dgCMatrix), Seurat's expected orientation
X <- as(usc2.macrophages.adjusted$X, "CsparseMatrix")
rownames(X) <- rownames(usc2.macrophages.adjusted$obs)
colnames(X) <- rownames(usc2.macrophages.adjusted$var)
counts <- Matrix::t(X)

# Integrate into Seurat object
usc2.macrophages[["despotx"]] = Seurat::CreateAssayObject(counts)


#==============================================================================================
#==============================================================================================

# scPearson DimReduc on decontaminated object
DefaultAssay(usc2.macrophages) = "decontx"


# Usage of scPEarsonPCA for DimReduc
# ==================================

tc = Matrix::colSums(usc2.macrophages[["decontx"]]$counts)

# create custom column regarding sample_id & fov as it seems that i observed batch effect related to both cases...
usc2.macrophages$sample_id_fov = paste0(usc2.macrophages$sample_id, "-", usc2.macrophages$fov)

# compute gene frequency...
genefreq_batch = scPearsonPCA::gene_frequency(
  usc2.macrophages[["decontx"]]$counts, 
  obs = data.table(usc2.macrophages@meta.data)[,.(cell_id, sample_id_fov)], 
  cellid_colname = "cell_id", 
  batch_variable = c("sample_id_fov")
)

usc2.macrophages = FindVariableFeatures(usc2.macrophages, nfeatures = 2000)
hvgs <- VariableFeatures(usc2.macrophages, assay = "decontx")

# compute PCA & Bacth integration...
pcaobj_batch = sparse_quasipoisson_pca_seurat_batch(
  usc2.macrophages[["decontx"]]$counts[hvgs,],
  totalcounts = tc,
  grate = genefreq_batch[hvgs,],
  obs =data.table(usc2.macrophages@meta.data)[,.(cell_id, sample_id_fov)],
  batch_variable = "sample_id_fov",
  cellid_colname = "cell_id",
  scale.max = 10,
  do.scale = TRUE,
  do.center = TRUE,
  ncores = 8
)

# perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2.macrophages[["pca_decontx"]] <- pcaobj_batch$reduction.data
usc2.macrophages[["umap_decontx"]] <- umapobj$ump  
usc2.macrophages[["graph_decontx"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2.macrophages <- Seurat::FindClusters(
  usc2.macrophages, 
  resolution = 1, 
  graph = "graph_decontx",
  algorithm = 4
)
usc2.macrophages@meta.data$clusters_decontx <- factor(paste0(
  "G",
  usc2.macrophages@meta.data$graph_decontx_res.1
))


# Find Markers...
Idents(usc2.macrophages) = usc2.macrophages$clusters_despotx

p3.macrophages = DimPlot(
  usc2.macrophages, 
  group.by = "clusters_decontx",
  label = T,
  raster = T,
  label.size = 4,
  reduction = "umap_decontx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")


p1.macrophages
p2.macrophages
p3.macrophages

```


```{r macrophages_decontX}

# Perform DecontX
# ===============
counts_rna <- GetAssayData(usc2.macrophages, assay = "RNA", layer = "counts")
dcx <- celda::decontX(
  x    = as(counts_rna, "CsparseMatrix"),
  z    = as.character(usc2.macrophages$clusters_despotx),
  seed = 1
)
usc2.macrophages$decontX_contamination <- dcx$contamination
usc2.macrophages[["decontX"]] <- CreateAssay5Object(counts = dcx$decontXcounts)


# scPearson DimReduc on decontaminated object
DefaultAssay(usc2.macrophages) = "decontX"


# Usage of scPEarsonPCA for DimReduc
# ==================================
tc = Matrix::colSums(usc2.macrophages[["decontX"]]$counts)

# create custom column regarding sample_id & fov as it seems that i observed batch effect related to both cases...
usc2.macrophages$sample_id_fov = paste0(usc2.macrophages$sample_id, "-", usc2.macrophages$fov)

# compute gene frequency...
genefreq_batch = scPearsonPCA::gene_frequency(
  usc2.macrophages[["decontX"]]$counts, 
  obs = data.table(usc2.macrophages@meta.data)[,.(cell_id, sample_id_fov)], 
  cellid_colname = "cell_id", 
  batch_variable = c("sample_id_fov")
)

usc2.macrophages = FindVariableFeatures(usc2.macrophages, nfeatures = 6215)
hvgs <- VariableFeatures(usc2.macrophages, assay = "decontX")

# compute PCA & Bacth integration...
pcaobj_batch = sparse_quasipoisson_pca_seurat_batch(
  usc2.macrophages[["decontX"]]$counts[hvgs,],
  totalcounts = tc,
  grate = genefreq_batch[hvgs,],
  obs =data.table(usc2.macrophages@meta.data)[,.(cell_id, sample_id_fov)],
  batch_variable = "sample_id_fov",
  cellid_colname = "cell_id",
  scale.max = 10,
  do.scale = TRUE,
  do.center = TRUE,
  ncores = 8
)

# perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2.macrophages[["pearsonpca_decontx"]] <- pcaobj_batch$reduction.data
usc2.macrophages[["pearsonumap_decontx"]] <- umapobj$ump  
usc2.macrophages[["pearsongraph_decontx"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2.macrophages <- Seurat::FindClusters(
  usc2.macrophages, 
  resolution = 1, 
  graph = "pearsongraph_decontx",
  algorithm = 4
)
usc2.macrophages@meta.data$clusters_decontx <- factor(paste0(
  "G",
  usc2.macrophages@meta.data$pearsongraph_decontx_res.1
))


Idents(usc2.macrophages) = usc2.macrophages$clusters_decontx
usc2.macrophages <- NormalizeData(
  usc2.macrophages,
  assay = "RNA",
  normalization.method = "RC",
  verbose = T
)

# Find markers using Seurat FindAllMarkers function...
markers.despotx <- FindAllMarkers(
  usc2.macrophages,
  assay           = "RNA",
  only.pos        = FALSE,
  min.pct         = 0.05,
  logfc.threshold = 0.1,
  test.use        = "wilcox"
)
fwrite(markers.despotx, file="all.markers.macrophages.decontx.tsv", sep="\t")

# Find markers using Hieratype ...
markers.despotx.htp = HieraType::clusterwise_foldchange_metrics(
  usc2.macrophages[["decontX"]]$counts[hvgs,],
  totalcounts = Matrix::colSums(usc2.macrophages[["decontX"]]$counts),
  metadata = usc2.macrophages@meta.data, 
  cluster_column = "clusters_decontx")
fwrite(markers.despotx.htp, file="all.markers.macrophages.decontx.htp.tsv", sep="\t")


# Annotation...
# =============
usc2.macrophages$celltype_macro <- dplyr::case_when(
  usc2.macrophages$clusters_decontx == "G3" ~ "Macrophages",
  usc2.macrophages$clusters_decontx == "G7" ~ "Dendritic.cells",
  usc2.macrophages$clusters_decontx == "G5" ~ "Neutrophils.Monocytes",

  usc2.macrophages$clusters_decontx == "G2" ~ "Contaminant.1",  # hepatocyte
  usc2.macrophages$clusters_decontx == "G6" ~ "Contaminant.2",  # fibroblast
  usc2.macrophages$clusters_decontx == "G8" ~ "Contaminant.3",  # plasma cells

  usc2.macrophages$clusters_decontx == "G1" ~ "Undefined.1",
  usc2.macrophages$clusters_decontx == "G4" ~ "Undefined.2",

  TRUE ~ "Undefined"
)

celltype_colors_macro <- c(
  "Macrophages"           = "#E6550D",
  "Dendritic.cells"       = "#31A354",
  "Neutrophils.Monocytes" = "#CB181D",

  "Contaminant.1" = "#D9D9D9",
  "Contaminant.2" = "#BDBDBD",
  "Contaminant.3" = "#A3A3A3",

  "Undefined.1" = "#404040",
  "Undefined.2" = "#2A2A2A",
  "Undefined"   = "#000000"
)


# Plotting...
# ===========

DimPlot(
  usc2.macrophages, 
  group.by = "clusters_decontx",
  label = T,
  label.size = 4,
  reduction = "pearsonumap_decontx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DecontX") +
  theme(legend.position = "none")

 DimPlot(
  usc2.macrophages, 
  group.by = "celltype_macro",
  label = T,
  label.size = 4,
  cols = celltype_colors_macro,
  repel = T,
  reduction = "pearsonumap_decontx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")
 
FeaturePlot(
  subset(usc2.macrophages, decontX_contamination < 0.7),
  order = T,
  label = T,
  raster = F,
  reduction = "pearsonumap_decontx",
  features = c("decontX_contamination"),
  assay = "decontX"
)


```

#### - b.2 T/NK.cells

```{r subset_Tcells, cache=T}

# subset Tcells
usc2.tcells = subset(usc2, celltype_despotx %in% c("T/NK.cells"))

# Perform Dimensionality reduction
# ================================

DefaultAssay(usc2.tcells) = "despotx"

# Usage of scPEarsonPCA for DimReduc
# ==================================

tc = Matrix::colSums(usc2.tcells[["despotx"]]$counts)

# create custom column regarding sample_id & fov as it seems that i observed batch effect related to both cases...
usc2.tcells$sample_id_fov = paste0(usc2.tcells$sample_id, "-", usc2.tcells$fov)

# compute gene frequency...
genefreq_batch = scPearsonPCA::gene_frequency(
  usc2.tcells[["despotx"]]$counts, 
  obs = data.table(usc2.tcells@meta.data)[,.(cell_id, sample_id_fov)], 
  cellid_colname = "cell_id", 
  batch_variable = c("sample_id_fov")
)

usc2.tcells = FindVariableFeatures(usc2.tcells, nfeatures = 500)
hvgs <- VariableFeatures(usc2.tcells, assay = "despotx")

# compute PCA & Bacth integration...
pcaobj_batch = sparse_quasipoisson_pca_seurat_batch(
  usc2.tcells[["despotx"]]$counts[hvgs,],
  totalcounts = tc,
  grate = genefreq_batch[hvgs,],
  obs =data.table(usc2.tcells@meta.data)[,.(cell_id, sample_id_fov)],
  batch_variable = "sample_id_fov",
  cellid_colname = "cell_id",
  scale.max = 10,
  do.scale = TRUE,
  do.center = TRUE,
  ncores = 8
)

# perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2.tcells[["pearsonpca"]] <- pcaobj_batch$reduction.data
usc2.tcells[["pearsonumap"]] <- umapobj$ump  
usc2.tcells[["pearsongraph"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2.tcells <- Seurat::FindClusters(
  usc2.tcells, 
  resolution = 1, 
  graph = "pearsongraph",
  algorithm = 4
)
usc2.tcells@meta.data$clusters <- factor(paste0(
  "G",
  usc2.tcells@meta.data$pearsongraph_res.1
))

DimPlot(
  usc2.tcells, 
  group.by = "celltype_despotx",
  label = T,
  label.size = 4,
  raster = T,
  repel = T,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")

DimPlot(
  usc2.tcells, 
  group.by = "clusters",
  label = T,
  label.size = 4,
  raster =T,
  repel = T,
  reduction = "pearsonumap") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction - DespotX") +
  theme(legend.position = "none")



# Conversion of Seurat object into a h5ad object / file
# =====================================================
tc = counts_mat <- Matrix::t(usc2.tcells[["RNA"]]$counts)

# 2. Cell metadata — the column DeSpotX will condition on via --cell-type-col.
obs_df <- data.frame(
  cluster = as.character(usc2.tcells$clusters),
  row.names = colnames(usc2.tcells)
)

# 3. Gene metadata — var_names only strictly required by DeSpotX.
var_df <- data.frame(row.names = rownames(usc2.tcells[["RNA"]]))

# 4. Spatial coordinates — must be EXACTLY 2D (x, y).
coords <- GetTissueCoordinates(usc2.tcells, image = "TA649")
coords <- coords[match(colnames(usc2.tcells), coords$cell), c("x", "y")]
spatial_mat <- as.matrix(coords)

# 5. Assemble....
usc2.tcells.adata <- make_anndata(
  X    = counts_mat,
  obs  = obs_df,
  var  = var_df,
  obsm = list(spatial = spatial_mat)
)

# 6. Force CSR before writing — via the raw Python object, not the R6 $X<- setter
py_ad <- usc2.tcells.adata$.get_py_object()
sp <- reticulate::import("scipy.sparse", convert = FALSE)
py_ad$X <- sp$csr_matrix(py_ad$X)

# sanity check before writing
cat("X format:", as.character(py_ad$X$format), "\n")   # should print "csr"
cat("X shape:", paste(reticulate::py_to_r(py_ad$X$shape), collapse=" x "), "\n")

# 7. Write directly from the raw python object (bypasses the R6 wrapper entirely).
py_ad$write_h5ad("usc2.tcells.h5ad")


```



















Two complementary approaches: **DecontX** (R / `celda`, cluster-based, CPU) run
inline here.

### 8.1 DecontX (celda, cluster-based)


```{r decontx, fig.width=5, fig.height=4, cache=T}

counts_rna <- GetAssayData(usc2, assay = "RNA", layer = "counts")   # genes x cells
dcx <- celda::decontX(
  x    = as(counts_rna, "CsparseMatrix"),
  z    = as.character(usc2$clusters),                                # cluster anchors
  seed = 1
)

usc2$decontX_contamination <- dcx$contamination
usc2[["decontX"]] <- CreateAssay5Object(counts = dcx$decontXcounts)
print(round(tapply(usc2$decontX_contamination, usc2$clusters, mean), 3))

#delete tmp_variables....
rm(counts_rna)
rm(dcx)

# contamination on the Harmony UMAP + per-cluster distribution
FeaturePlot(
  usc2, "decontX_contamination", 
  reduction = "pearsonumap",
  assay = "decontX") +
  ggtitle("DecontX contamination")

ggplot(
  usc2@meta.data, aes(clusters, decontX_contamination, fill = clusters)) +
  geom_boxplot(outlier.size = 0.1) + theme_clean(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(x = NULL, y = "contamination", title = "Contamination by cluster")

```

Since contamination is moderate and uneven across clusters,
we redo the PCA → Harmony → UMAP → clustering pipeline
on the `decontX` assay.

```{r decontx_reembed, fig.width=5, fig.height=4, cache=T}

DefaultAssay(usc2) = "decontX"

tc <- Matrix::colSums(usc2[["decontX"]]$counts)
genefreq_batch <- 
  gene_frequency(usc2[["decontX"]]$counts, obs = data.table(usc2@meta.data)[,.(cell_id, sample_id)]
                 ,cellid_colname = "cell_id"
                 ,batch_variable = "sample_id")

usc2 = FindVariableFeatures(usc2, nfeatures = 700)
hvgs <- VariableFeatures(usc2, assay = "RNA")

### Returns a Seurat-style DimReduc object with 
pcaobj_batch <- 
  sparse_quasipoisson_pca_seurat_batch(usc2[["decontX"]]$counts[hvgs,]
                                       ,totalcounts = tc
                                       ,grate = genefreq_batch[hvgs,]
                                       ,obs =data.table(usc2@meta.data)[,.(cell_id, sample_id)] 
                                       ,batch_variable = "sample_id"
                                       ,cellid_colname = "cell_id"
                                       ,scale.max = 10
                                       ,do.scale = TRUE
                                       ,do.center = TRUE,
                                       ncores = 8
  )
#perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2[["pearsonpca_decontX"]] <- pcaobj_batch$reduction.data
usc2[["pearsonumap_decontX"]] <- umapobj$ump  
usc2[["pearsongraph_decontX"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2 <- Seurat::FindClusters(
  usc2,
  resolution = 1,
  algorithm = 4,
  graph = "pearsongraph_decontX")
usc2@meta.data$clusters_decontx <- paste0(
  "G",
  usc2@meta.data$pearsongraph_res.1)

# Plotting...
DimPlot(
  usc2, 
  group.by = "sample_id",
  label = F,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

DimPlot(
  usc2, 
  group.by = "clusters_decontx",
  label = T,
  raster = T,
  label.size = 3,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — RNA, scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```


```{r scPearson_decontX_markers, fig.width=5, fig.height=10, cache=T}

# Set default reductions
Idents(usc2) = usc2$clusters_decontx

usc2 <- NormalizeData(
  usc2,
  assay = "RNA",
  normalization.method = "RC",
  verbose = T
)

markers_decontX <- FindAllMarkers(
  usc2,
  assay           = "RNA",
  only.pos        = FALSE,
  min.pct         = 0.05,
  logfc.threshold = 0.1,
  test.use        = "wilcox"
)

markers_reliable <- markers_decontX |>
  dplyr::filter(
    p_val_adj < 0.05,
    abs(avg_log2FC) >= 0.25,
    abs(pct.1 - pct.2) >= 0.10
  )

#write tables
fwrite(markers_decontX, file = "markers_decontX.tsv", sep="\t")

top_markers = markers_reliable %>%
  group_by(cluster) %>%
  slice_head(n=10) %>%
  data.table()

# show Heatmap of markers
markers.mat = Seurat::AverageExpression(
  usc2,
  features = top_markers$gene %>% unique(),
  assay = "RNA", slot = "counts"
)$RNA %>% data.frame()

pheatmap::pheatmap(markers.mat, scale="row", fontsize_row = 1)

```

Let's look the expression of canonical markers in the data


```{r hierratype_decontX, fig.width=15, fig.height=7, cache=T}

fctbl_rna <- clusterwise_foldchange_metrics(usc2[["decontX"]]$counts[hvgs,]
                                            ,totalcounts = tc
                                            ,metadata = usc2@meta.data
                                            ,cluster_column = "celltype_decontX"
                                            )

evidence_genes <- list(
  "Hepatocytes.1" = c("FASN","ACSL5","MSMO1","AQP9","TBX3","GJB1"),
  "Hepatocytes.2" = c("PLA2G2A","CHI3L1","SERPINE1","CREB3L3","GHR","LSR"),
  "Hepatocytes.3" = c("GCGR","HLF","KLKB1","PROX1"),
  "Hepatocytes.4" = c("CXCL9","CXCL10","GBP1","GBP4","GBP5","STAT1","ISG15"),
  "Hepatocytes.5" = c("BCHE","TFPI","CCL16","COX1","COX2"),
  "Macrophages"   = c("MARCO","CD5L","CD163","VSIG4","C1QA","C1QB","C1QC",
                       "GPNMB","LGALS3","HLA-DQA1","HLA-DRA"),
  "Fibroblasts.1" = c("COL1A1","COL3A1","ACTA2","TAGLN","PDGFRB"),
  "Fibroblasts.2" = c("RELN","HGF","AGTR1","RGS5"),
  "Endothelials"  = c("STAB1","KDR","CLDN5","CDH5","LYVE1"),
  "T/NK.cells"    = c("CD8B","GNLY","GZMA","TRBC1","CD27"),
  "Plasma.cells"  = c("IGKC","IGLC1/2","IGLL5","JCHAIN","MZB1","POU2AF1"),
  "Cholangiocytes"= c("KRT7","KRT19","EPCAM","SOX9","CDH1"),
  "Proliferating" = c("TOP2A","CCNB1","PCNA","STMN1"),
  "Monocytes.Neutrophils" = c("S100A8","S100A9","FCN1","CSF3R","NCF1")
)

fctbl_rna = fctbl_rna %>% filter(gene %in% unique(unlist(evidence_genes)))
hm_rna <- add_ncells(marker_heatmap(fctbl_rna)) + labs(title = "rna heatmap")
print(hm_rna)

```


Seems reaaalyy good ...Let's proceed now with annotation...

```{r decontX_annotation, fig.width=5, fig.height=4, cache=T}

usc2$celltype_decontX <- dplyr::case_when(
  usc2$clusters_decontx == "G1"  ~ "Hepatocytes.1",   # lipogenic (FASN, ACSL5, MSMO1)
  usc2$clusters_decontx == "G4"  ~ "Hepatocytes.2",   # acute-phase/stress (SERPINE1, CREB3L3)
  usc2$clusters_decontx == "G5"  ~ "Hepatocytes.3",   # periportal/gluconeogenic (GCGR, HLF)
  usc2$clusters_decontx == "G9"  ~ "Hepatocytes.4",   # IFN-γ-stimulated (CXCL9/10, GBP1-5, STAT1)
  usc2$clusters_decontx == "G13" ~ "Hepatocytes.5",   # core/purest hepatocyte (BCHE, TFPI, CCL16)

  usc2$clusters_decontx == "G3"  ~ "Macrophages",     # Kupffer + scar-associated merged this round

  usc2$clusters_decontx == "G6"  ~ "Fibroblasts.1",   # activated myofibroblast (COL1A1, ACTA2)
  usc2$clusters_decontx == "G7"  ~ "Fibroblasts.2",   # quiescent HSC/portal (RELN, HGF, AGTR1)

  usc2$clusters_decontx == "G10" ~ "Endothelials",    # LSEC (STAB1, KDR, CLDN5, LYVE1)

  usc2$clusters_decontx == "G2"  ~ "T/NK.cells",

  usc2$clusters_decontx == "G8"  ~ "Plasma.cells",    # kappa + lambda merged this round

  usc2$clusters_decontx == "G11" ~ "Cholangiocytes",

  usc2$clusters_decontx == "G12" ~ "Proliferating",   # pure cell-cycle signature, lineage-ambiguous

  usc2$clusters_decontx == "G14" ~ "Monocytes/Neutrophils",

  TRUE ~ "Undefined"
)

celltype_colors <- c(
  # Hepatocytes — blue shades (5 substates)
  "Hepatocytes.1" = "#08306B",
  "Hepatocytes.2" = "#08519C",
  "Hepatocytes.3" = "#2171B5",
  "Hepatocytes.4" = "#4292C6",
  "Hepatocytes.5" = "#6BAED6",

  "Macrophages"   = "#E6550D",

  # Fibroblasts — purple shades
  "Fibroblasts.1" = "#6A3D9A",
  "Fibroblasts.2" = "#A27AC3",

  # Other major lineages
  "Endothelials"          = "#00A087",
  "T/NK.cells"            = "#33A02C",
  "Plasma.cells"          = "#C51B7D",
  "Cholangiocytes"        = "#D8A600",
  "Proliferating"         = "#525252",
  "Monocytes/Neutrophils" = "#CB181D",
  "Undefined"             = "#BDBDBD"
)


DimPlot(
  usc2, 
  group.by = "clusters_decontx",
  label = T,
  raster = T,
  label.size = 4,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")


DimPlot(
  usc2, 
  group.by = "celltype_decontX",
  cols = celltype_colors,
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```

In order to better characterize T.cells subtypes ....  I select immunes types and reperform dim
reduc analysis


```{r immune_selection, fig.width=5, fig.height=4, cache=T}

usc2_immune = subset(
  usc2,
  (clusters %in% c("G14","G3"))
)

DefaultAssay(usc2_immune) = "RNA"

DimPlot(
  usc2_immune, 
  group.by = "celltype_decontX",
  cols = celltype_colors,
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```


```{r immune_scPearson_dimReduc, fig.width=4, fig.height=4, cache=T}

tc <- Matrix::colSums(usc2_immune[["RNA"]]$counts)
genefreq_batch <- 
  gene_frequency(usc2_immune[["RNA"]]$counts, obs = data.table(usc2_immune@meta.data)[,.(cell_id, sample_id)]
                 ,cellid_colname = "cell_id"
                 ,batch_variable = "sample_id")

usc2_immune = FindVariableFeatures(usc2_immune, nfeatures = 6215)
hvgs <- VariableFeatures(usc2_immune, assay = "RNA")

### Returns a Seurat-style DimReduc object with 
pcaobj_batch <- 
  sparse_quasipoisson_pca_seurat_batch(usc2_immune[["RNA"]]$counts[hvgs,]
                                       ,totalcounts = tc
                                       ,grate = genefreq_batch[hvgs,]
                                       ,obs =data.table(usc2_immune@meta.data)[,.(cell_id, sample_id)] 
                                       ,batch_variable = "sample_id"
                                       ,cellid_colname = "cell_id"
                                       ,scale.max = 10
                                       ,do.scale = TRUE
                                       ,do.center = TRUE
  )
#perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2_immune[["pearsonpca"]] <- pcaobj_batch$reduction.data
usc2_immune[["pearsonumap"]] <- umapobj$ump  
usc2_immune[["pearsongraph"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2_immune <- Seurat::FindClusters(usc2_immune, resolution = 1.5, graph = "pearsongraph", algorithm = 4)
usc2_immune@meta.data$clusters <- usc2_immune@meta.data$pearsongraph_res.1.5

DimPlot(
  usc2_immune, 
  group.by = "sample_id",
  label = F,
  label.size = 4,
  raster = T,
  reduction = "pearsonumap") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages cells — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

DimPlot(
  usc2_immune, 
  group.by = "clusters",
  label = T,
  label.size = 4,
  raster = T,
  reduction = "pearsonumap") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages cells — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

```

Let's perform now another round of decontamintaion of this subset
This time I will use on the tools suggested by Vivek that I think perform way better than decontX
to remove contamination related to the hepatocytes,

```{r immune_decontX,  fig.width=4, fig.height=4, cache=T}

counts_rna <- GetAssayData(usc2_immune, assay = "RNA", layer = "counts")   # genes x cells
dcx <- celda::decontX(
  x    = as(counts_rna, "CsparseMatrix"),
  z    = as.character(usc2_immune$clusters),                                # cluster anchors
  seed = 1
)

usc2_immune$decontX_contamination <- dcx$contamination
usc2_immune[["decontX"]] <- CreateAssay5Object(counts = dcx$decontXcounts)
print(round(tapply(usc2_immune$decontX_contamination, usc2_immune$clusters, mean), 3))

#delete tmp_variables....
rm(counts_rna)
rm(dcx)

# contamination on the Harmony UMAP + per-cluster distribution
FeaturePlot(
  usc2_immune, "decontX_contamination", 
  reduction = "pearsonumap",
  assay = "decontX") +
  ggtitle("DecontX contamination")

ggplot(
  usc2_immune@meta.data, aes(clusters, decontX_contamination, fill = clusters)) +
  geom_boxplot(outlier.size = 0.1) + theme_clean(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(x = NULL, y = "contamination", title = "Contamination by cluster")


```

```{r immune_dimreduc_decontX, fig.width=5, fig.height=4, cache=T}

DefaultAssay(usc2_immune) = "decontX"

tc <- Matrix::colSums(usc2_immune[["decontX"]]$counts)
genefreq_batch <- 
gene_frequency(usc2_immune[["decontX"]]$counts, obs = data.table(usc2_immune@meta.data)[,.(cell_id, sample_id)]
               ,cellid_colname = "cell_id"
               ,batch_variable = "sample_id")

usc2_immune = FindVariableFeatures(usc2_immune, nfeatures = 6215)
hvgs <- VariableFeatures(usc2_immune, assay = "RNA")

### Returns a Seurat-style DimReduc object with 
pcaobj_batch <- 
sparse_quasipoisson_pca_seurat_batch(usc2_immune[["decontX"]]$counts[hvgs,]
                                     ,totalcounts = tc
                                     ,grate = genefreq_batch[hvgs,]
                                     ,obs =data.table(usc2_immune@meta.data)[,.(cell_id, sample_id)] 
                                     ,batch_variable = "sample_id"
                                     ,cellid_colname = "cell_id"
                                     ,scale.max = 10
                                     ,do.scale = TRUE
                                     ,do.center = TRUE,
                                     ncores = 8
                                     )
#perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2_immune[["pearsonpca_decontX"]] <- pcaobj_batch$reduction.data
usc2_immune[["pearsonumap_decontX"]] <- umapobj$ump  
usc2_immune[["pearsongraph_decontX"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2_immune <- Seurat::FindClusters(usc2_immune, resolution = 1.5, graph = "pearsongraph_decontX", algorithm = 4, verbose = T)
usc2_immune@meta.data$clusters_macro <- paste0(
  "G",
  usc2_immune@meta.data$pearsongraph_decontX_res.1.5)

DimPlot(
  usc2_immune, 
  group.by = "sample_id",
  label = F,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

DimPlot(
  usc2_immune, 
  group.by = "clusters_macro",
  label = T,
  raster = T,
  label.size = 3,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```

```{r immune_dimreduc_decontX_markers,  fig.width=4, fig.height=9, cache=T}

# Set default reductions
Idents(usc2_immune) = usc2_immune$clusters_macro

usc2_immune <- NormalizeData(
  usc2_immune,
  assay = "RNA",
  normalization.method = "RC",
  verbose = FALSE
)

markers_decontX <- FindAllMarkers(
  usc2_immune,
  assay           = "RNA",
  only.pos        = FALSE,
  min.pct         = 0.05,
  logfc.threshold = 0.05,
  test.use        = "wilcox"
)

markers_reliable <- markers_decontX |>
  dplyr::filter(
    p_val_adj < 0.05,
    abs(avg_log2FC) >= 0.25,
    abs(pct.1 - pct.2) >= 0.10
)

#write tables
fwrite(markers_decontX, file = "markers_decontX_immune.tsv", sep="\t")

# top_markers = markers_reliable %>%
#   group_by(cluster) %>%
#   slice_head(n=10) %>%
#   data.table()
# 
# # show Heatmap of markers
# markers.mat = Seurat::AverageExpression(
#   usc2_immune,
#   features = top_markers$gene %>% unique(),
#   assay = "RNA", slot = "counts"
#   )$RNA %>% data.frame()

# pheatmap::pheatmap(markers.mat, scale="row", fontsize_row = 5)

```


```{r immune_annotation, fig.width=5, fig.height=5, cache=T}
usc2_immune$celltype_immune <- dplyr::case_when(
  usc2_immune$clusters_macro == "G3"  ~ "Macrophages/Kupffer.1",   # Kupffer, resident/core — VSIG4+MARCO+CD5L+FOLR2+CD163
  usc2_immune$clusters_macro == "G9"  ~ "Macrophages/Kupffer.2",   # Kupffer-related, MRC1-high (KC2-leaning, FOLR2-negative)
  usc2_immune$clusters_macro == "G13" ~ "Macrophages/SAM",   # scar-associated (SAM) — GPNMB+LGALS3+FABP5+CD9+CD63

  usc2_immune$clusters_macro == "G7"  ~ "Dendritic.cells",     # CIITA/IRF8/IDO1, pan-macrophage core negative
  usc2_immune$clusters_macro == "G8"  ~ "Recruited.myeloid",   # VCAN+FCN1+S100A8/9+CXCR1/2, pan-macrophage strongly negative

  usc2_immune$clusters_macro == "G1"  ~ "Contaminant.1",    # Hepatocytes
  usc2_immune$clusters_macro == "G6"  ~ "Contaminant.2",    # Fibroblasts (weak SAM/macrophage bleed — likely real scar-niche admixture, not noise)
  usc2_immune$clusters_macro == "G10" ~ "Contaminant.3",    # T/NK cells
  usc2_immune$clusters_macro == "G12" ~ "Contaminant.4",    # Plasma cells
  usc2_immune$clusters_macro == "G11" ~ "Contaminant.5",    # Hepatocytes, stressed/heat-shock overlay

  usc2_immune$clusters_macro == "G2"  ~ "Undefined.1",
  usc2_immune$clusters_macro == "G4"  ~ "Undefined.2",
  usc2_immune$clusters_macro == "G5"  ~ "Undefined.3",

  TRUE ~ "Undefined"
)

celltype_colors <- c(
  "Macrophages/Kupffer.1"     = "#E6550D",
  "Macrophages/Kupffer.2"     = "#A63603",
  "Macrophages/SAM"     = "#FD8D3C",

  "Dendritic.cells"   = "#31A354",
  "Recruited.myeloid" = "#8C510A",

  "Contaminant.1" = "#F0F0F0",
  "Contaminant.2" = "#D9D9D9",
  "Contaminant.3" = "#BDBDBD",
  "Contaminant.4" = "#A3A3A3",
  "Contaminant.5" = "#898989",

  "Undefined.1" = "#404040",
  "Undefined.2" = "#2A2A2A",
  "Undefined.3" = "#151515",
  "Undefined"   = "#000000"
)

DimPlot(
  usc2_immune, 
  group.by = "clusters_macro",
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

DimPlot(
  usc2_immune, 
  group.by = "celltype_immune",
  cols = celltype_colors,
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```

```{r hieratype_macrophages, fig.width=10, fig.height=5, cache=T}


fctbl_rna <- clusterwise_foldchange_metrics(usc2_immune[["decontX"]]$counts[hvgs,]
                                            ,totalcounts = tc
                                            ,metadata = usc2_immune@meta.data
                                            ,cluster_column = "celltype_immune"
                                            )


evidence_genes <- list(
  "Macrophages.1" = c("VSIG4","MARCO","CD5L","FOLR2","SLC40A1","HMOX1","CD163"),
  # Kupffer, resident/core — FOLR2+ is what separates this from Macrophages.2

  "Macrophages.2" = c("MRC1","MARCO","CD5L","CD163","HMOX1","VSIG4"),
  # Kupffer-related, MRC1-high/KC2-leaning — same core minus FOLR2, MRC1 is the standout

  "Macrophages.3" = c("GPNMB","LGALS3","CD9","FABP5","CD63"),
  # scar-associated (SAM) — GPNMB/LGALS3/FABP5 axis specifically

  "Dendritic.cells" = c("CIITA","IRF8","IDO1","HLA-DMA","HLA-DQA1","HLA-DQB1",
                         "HLA-DPA1","HLA-DPB1","HLA-DRA","HLA-DRB","ITGAX"),
  # pan-macrophage core (CD68/C1QA-C/CD163) is NEGATIVE here — this is what distinguishes it from macrophages

  "Recruited.myeloid" = c("VCAN","FCN1","S100A8","S100A9","CXCR1","CXCR2","CD14","CSF3R")
  # neutrophil + classical monocyte mixed — pan-macrophage core strongly negative
)

fctbl_rna = fctbl_rna %>% filter(gene %in% unique(unlist(evidence_genes)))

fctbl_rna$cluster = factor(
  fctbl_rna$cluster,
  levels = c("Macrophages/Kupffer.1", "Macrophages/Kupffer.2", "Dendritic.cells", "Macrophages/SAM",
             "Recruited.myeloid", "Contaminant.2", "Contaminant.4", "Contaminant.5",
             "Undefined.1", "Undefined.2", "Contaminant.3", "Undefined.3", "Contaminant.1")
)

hm_rna <- add_ncells(marker_heatmap(fctbl_rna)) + labs(title = "rna heatmap")
print(hm_rna)

```



```{r tcells_selection, fig.width=4, fig.height=4, cache=T}

usc2_tcells = subset(
  usc2,
  (clusters %in% c("G2"))
)

DefaultAssay(usc2_tcells) = "RNA"

DimPlot(
  usc2_tcells, 
  group.by = "celltype_decontX",
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```

```{r tcells_scPearson_dimReduc, fig.width=4, fig.height=4, cache=T}

tc <- Matrix::colSums(usc2_tcells[["RNA"]]$counts)
genefreq_batch <- 
gene_frequency(usc2_tcells[["RNA"]]$counts, obs = data.table(usc2_tcells@meta.data)[,.(cell_id, sample_id)]
               ,cellid_colname = "cell_id"
               ,batch_variable = "sample_id")

usc2_tcells = FindVariableFeatures(usc2_tcells, nfeatures = 6125)
hvgs <- VariableFeatures(usc2_tcells, assay = "RNA")

### Returns a Seurat-style DimReduc object with 
pcaobj_batch <- 
sparse_quasipoisson_pca_seurat_batch(usc2_tcells[["RNA"]]$counts[hvgs,]
                                     ,totalcounts = tc
                                     ,grate = genefreq_batch[hvgs,]
                                     ,obs =data.table(usc2_tcells@meta.data)[,.(cell_id, sample_id)] 
                                     ,batch_variable = "sample_id"
                                     ,cellid_colname = "cell_id"
                                     ,scale.max = 10
                                     ,do.scale = TRUE
                                     ,do.center = TRUE,
                                     ncores = 8
                                     )
#perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2_tcells[["pearsonpca"]] <- pcaobj_batch$reduction.data
usc2_tcells[["pearsonumap"]] <- umapobj$ump  
usc2_tcells[["pearsongraph"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2_tcells <- Seurat::FindClusters(usc2_tcells, resolution = 1.4, graph = "pearsongraph")
usc2_tcells@meta.data$clusters_tcells <- usc2_tcells@meta.data$pearsongraph_res.1.4

DimPlot(
  usc2_tcells, 
  group.by = "sample_id",
  label = F,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells  — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

DimPlot(
  usc2_tcells, 
  group.by = "clusters_tcells",
  label = T,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells  — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

```

Let's perform now another round of decontamintaion of this subset

```{r immune_decontX,  fig.width=4, fig.height=4, cache=T}

counts_rna <- GetAssayData(usc2_tcells, assay = "RNA", layer = "counts")   # genes x cells
dcx <- celda::decontX(
  x    = as(counts_rna, "CsparseMatrix"),
  z    = as.character(usc2_tcells$clusters_tcells),                                # cluster anchors
  seed = 1
)

usc2_tcells$decontX_contamination <- dcx$contamination
usc2_tcells[["decontX"]] <- CreateAssay5Object(counts = dcx$decontXcounts)

#delete tmp_variables....
rm(counts_rna)
rm(dcx)

# contamination on the Harmony UMAP + per-cluster distribution
FeaturePlot(
  usc2_tcells, "decontX_contamination", 
  reduction = "pearsonumap",
  assay = "decontX") +
  ggtitle("DecontX contamination")

ggplot(
  usc2_tcells@meta.data, aes(clusters_tcells, decontX_contamination, fill = clusters_tcells)) +
  geom_boxplot(outlier.size = 0.1) + theme_clean(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(x = NULL, y = "contamination", title = "Contamination by cluster")


```


```{r tcells_dimreduc_decontX, fig.width=4, fig.height=4, cache=T}

DefaultAssay(usc2_tcells) = "decontX"

tc <- Matrix::colSums(usc2_tcells[["decontX"]]$counts)
genefreq_batch <- 
  gene_frequency(usc2_tcells[["decontX"]]$counts, obs = data.table(usc2_tcells@meta.data)[,.(cell_id, sample_id)]
                 ,cellid_colname = "cell_id"
                 ,batch_variable = "sample_id")

usc2_tcells = FindVariableFeatures(usc2_tcells, nfeatures = 6215)
hvgs <- VariableFeatures(usc2_tcells, assay = "RNA")

### Returns a Seurat-style DimReduc object with 
pcaobj_batch <- 
  sparse_quasipoisson_pca_seurat_batch(usc2_tcells[["decontX"]]$counts[hvgs,]
                                       ,totalcounts = tc
                                       ,grate = genefreq_batch[hvgs,]
                                       ,obs =data.table(usc2_tcells@meta.data)[,.(cell_id, sample_id)] 
                                       ,batch_variable = "sample_id"
                                       ,cellid_colname = "cell_id"
                                       ,scale.max = 10
                                       ,do.scale = TRUE
                                       ,do.center = TRUE,
                                       ncores = 8
  )
#perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc2_tcells[["pearsonpca_decontX"]] <- pcaobj_batch$reduction.data
usc2_tcells[["pearsonumap_decontX"]] <- umapobj$ump  
usc2_tcells[["pearsongraph_decontX"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc2_tcells <- Seurat::FindClusters(usc2_tcells, resolution = 2, graph = "pearsongraph_decontX", algorithm = 4)
usc2_tcells@meta.data$clusters_tcells_decontx <- paste0(
  "G",
  usc2_tcells@meta.data$pearsongraph_decontX_res.2)

DimPlot(
  usc2_tcells, 
  group.by = "sample_id",
  label = F,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

DimPlot(
  usc2_tcells, 
  group.by = "clusters_tcells_decontx",
  label = T,
  raster = T,
  label.size = 3,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```


```{r tcells_markers,  fig.width=4, fig.height=9, cache=T}

# Set default reductions
Idents(usc2_tcells) = usc2_tcells$clusters_tcells_decontx

usc2_tcells <- NormalizeData(
  usc2_tcells,
  assay = "RNA",
  normalization.method = "RC",
)

markers_decontX <- FindAllMarkers(
  usc2_tcells,
  assay           = "RNA",
  only.pos        = FALSE,
  min.pct         = 0.01,
  logfc.threshold = 0.05,
  test.use        = "wilcox"
)

markers_reliable <- markers_decontX |>
  dplyr::filter(
    p_val_adj < 0.05,
    abs(avg_log2FC) >= 0.25,
    abs(pct.1 - pct.2) >= 0.10
  )

#write tables
fwrite(markers_decontX, file = "markers_decontX_tcells.tsv", sep="\t")


# top_markers = markers_reliable %>%
#   group_by(cluster) %>%
#   slice_head(n=30) %>%
#   data.table()
# 
# # show Heatmap of markers
# markers.mat = Seurat::AverageExpression(
#   usc2_tcells,
#   features = top_markers$gene %>% unique(),
#   assay = "RNA", slot = "counts"
# )$RNA %>% data.frame()
# 
# pheatmap::pheatmap(markers.mat, scale="row", fontsize_row = 5)

```



```{r hieratype_tcells,  fig.width=10, fig.height=5, cache=T}

data("markerslist_tcellmajor")
data(pipeline_tcell)

fct <- clusterwise_foldchange_metrics(counts = usc2_tcells[["decontX"]]$counts[hvgs,],
                                      ,metadata = usc2_tcells@meta.data,
                                      ,cluster_column = "clusters_tcells_decontx")
idxs <- unique(unname(c(unlist(lapply(HieraType::markerslist_cd4tminor, "[[", "index_marker"))
               ,unlist(lapply(HieraType::markerslist_cd8tminor, "[[", "index_marker")))))

nms <- c("cd4t", names(HieraType::markerslist_cd4tminor), "cd8t", names(HieraType::markerslist_cd8tminor))
hmsubtype <- marker_heatmap(
  fct, 
  featsuse = unique(c("CD8A", "CD8B", "CD4", idxs)), 
  fold_change_min = 0.05, 
  cluster_prop_min = 0.05, 
  topleft_to_bottomright = F) ;print(hmsubtype)


```




```{r tcells_annotation, fig.width=6, fig.height=4, cache=T}

usc2_tcells$celltype_tcell <- dplyr::case_when(
  usc2_tcells$clusters_tcells_decontx == "G5"  ~ "cd8_tem_cytotoxic",   # fold_change 2.50/1.95 — clear signal
  usc2_tcells$clusters_tcells_decontx == "G3"  ~ "cd4_treg",            # 1.48
  usc2_tcells$clusters_tcells_decontx == "G8"  ~ "cd4_cd8_tcm",         # 1.74
  usc2_tcells$clusters_tcells_decontx == "G13" ~ "cd8_exhausted",       # 1.50, but only 64 cells — tentative
  usc2_tcells$clusters_tcells_decontx == "G6"  ~ "naive_like",          # 1.33/1.30, cd4/cd8 tied
  usc2_tcells$clusters_tcells_decontx == "G9"  ~ "cd8_exhausted_cytotoxic",  # 1.31/1.27, weak
  usc2_tcells$clusters_tcells_decontx == "G4"  ~ "cd4_th2",             # 1.29, weak
  usc2_tcells$clusters_tcells_decontx == "G11" ~ "cd4_tem",             # 1.27, weak
  usc2_tcells$clusters_tcells_decontx == "G2"  ~ "cd4_th2_treg",        # 1.27/1.23, weak/ambiguous
  usc2_tcells$clusters_tcells_decontx == "G10" ~ "cd8_tem",             # 1.20, weak

  usc2_tcells$clusters_tcells_decontx == "G1"  ~ "Undefined.1",   # max fold_change 0.78 — no signal, matches contaminated UMAP island
  usc2_tcells$clusters_tcells_decontx == "G7"  ~ "Undefined.2",   # max fold_change 1.18 — no signal
  usc2_tcells$clusters_tcells_decontx == "G12" ~ "Undefined.3",   # max fold_change 1.07 — no signal

  TRUE ~ "Undefined"
)


celltype_colors_tcell <- c(
  # CD8 lineage — blue shades (darker = higher confidence)
  "cd8_tem_cytotoxic"       = "#08519C",  # G5, strong signal
  "cd8_exhausted"           = "#3182BD",  # G13, tentative (n=64)
  "cd8_exhausted_cytotoxic" = "#6BAED6",  # G9, weak
  "cd8_tem"                 = "#C6DBEF",  # G10, weak

  # CD4 lineage — green/purple shades
  "cd4_treg"    = "#6A3D9A",  # G3, moderate
  "cd4_th2"     = "#DE77AE",  # G4, weak
  "cd4_tem"     = "#A6D96A",  # G11, weak

  # mixed / shared / ambiguous calls
  "cd4_cd8_tcm"  = "#1B7837",  # G8, moderate
  "naive_like"   = "#41AB5D",  # G6, moderate
  "cd4_th2_treg" = "#C994C7",  # G2, weak/ambiguous

  "Undefined.1" = "#404040",
  "Undefined.2" = "#2A2A2A",
  "Undefined.3" = "#151515"
)


DimPlot(
  usc2_tcells, 
  group.by = "clusters_tcells_decontx",
  # cols = celltype_colors,
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

DimPlot(
  usc2_tcells, 
  group.by = "celltype_tcell",
  # cols = celltype_colors,
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```



```{r hieratype_ctypes_tcells, fig.width=10, fig.height=5, cache=T}


data("markerslist_tcellmajor")
data(pipeline_tcell)

fct <- clusterwise_foldchange_metrics(counts = usc2_tcells[["decontX"]]$counts[hvgs,],
                                      ,metadata = usc2_tcells@meta.data,
                                      ,cluster_column = "celltype_tcell")
idxs <- unique(unname(c(unlist(lapply(HieraType::markerslist_cd4tminor, "[[", "index_marker"))
               ,unlist(lapply(HieraType::markerslist_cd8tminor, "[[", "index_marker")))))

nms <- c("cd4t", names(HieraType::markerslist_cd4tminor), "cd8t", names(HieraType::markerslist_cd8tminor))
hmsubtype <- marker_heatmap(
  fct, 
  featsuse = unique(c("CD8A", "CD8B", "CD4", idxs)), 
  fold_change_min = 0.05, 
  cluster_prop_min = 0.05, 
  topleft_to_bottomright = F) ;print(hmsubtype)



```


## 5. DeSpotX-corrected counts → assay + corrected embedding

Load the DeSpotX-adjusted counts as a `despotx` assay on `usc` (same cells/order),
carry the per-cell `epsilon` contamination, then run the same scPearson
batch-corrected PCA/UMAP + clustering + markers as §3/§4.

```{r read_despotx}

suppressPackageStartupMessages({
  library(zellkonverter); library(SingleCellExperiment); library(SummarizedExperiment)
})

DESPOTX_H5AD = "~/Projects/TA649_analysis/adjusted.h5ad"


sce_dx <- readH5AD(DESPOTX_H5AD, reader = "R")

dx_counts <- as(assay(sce_dx, assayNames(sce_dx)[1]), "CsparseMatrix")   # genes x cells
dx_counts <- dx_counts[rownames(usc2[["RNA"]]), colnames(usc2)]           # align genes & cells

usc2[["despotx"]] <- CreateAssay5Object(counts = dx_counts)
usc2$despotx_epsilon <- as.data.frame(colData(sce_dx))[colnames(usc2), "epsilon"]

rm(sce_dx, dx_counts); gc()
```



```{r dimreduc_despotx,  fig.width=7, fig.height=7, cache=T}
DefaultAssay(usc2) <- "despotx"
usc3 <- NormalizeData(usc2, assay = "despotx", normalization.method = "RC")  # for markers

tc <- Matrix::colSums(usc3[["despotx"]]$counts)

genefreq_batch <- gene_frequency(
  usc3[["despotx"]]$counts,
  obs            = data.table(usc3@meta.data)[, .(cell_id, sample_id)],
  cellid_colname = "cell_id",
  batch_variable = "sample_id"
)

# Reuse the RNA HVG set (§3) so all three embeddings use the same genes
hvgs <- VariableFeatures(usc3, assay = "RNA")

pcaobj_batch <- sparse_quasipoisson_pca_seurat_batch(
  usc3[["despotx"]]$counts[hvgs, ],
  totalcounts    = tc,
  grate          = genefreq_batch[hvgs, ],
  obs            = data.table(usc3@meta.data)[, .(cell_id, sample_id)],
  batch_variable = "sample_id",
  cellid_colname = "cell_id",
  scale.max      = 10,
  do.scale       = TRUE,
  do.center      = TRUE,
  ncores = 8
)

umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

usc3[["pearsonpca_despotx"]]   <- pcaobj_batch$reduction.data
usc3[["pearsonumap_despotx"]]  <- umapobj$ump
usc3[["pearsongraph_despotx"]] <- Seurat::as.Graph(umapobj$grph)

usc3 <- Seurat::FindClusters(usc3, resolution = 1.5, graph = "pearsongraph_despotx", algorithm=4)
usc3@meta.data$clusters_despotx <- factor(paste0("G", usc3@meta.data$pearsongraph_despotx_res.1.5))

# Cluster markers (despotx). Same permissive thresholds; gate on adjusted p.
Idents(usc3) <- usc3$clusters_despotx

DimPlot(
  usc3, 
  group.by = "sample_id",
  label = F,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages scPearson Batch correction -> despotX applied") +
  theme(legend.position = "none")

DimPlot(
  usc3, 
  group.by = "clusters_despotx",
  label = T,
  raster = T,
  label.size = 3,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages scPearson Batch correction -> deSpotX applied") +
  theme(legend.position = "none")

```


```{r markers,  fig.width=4, fig.height=9, cache=T}

# Set default reductions
Idents(usc3) = usc3$clusters_despotx

usc3 <- NormalizeData(
  usc3,
  assay = "RNA",
  normalization.method = "RC",
)

markers_decontX <- FindAllMarkers(
  usc3,
  assay           = "RNA",
  only.pos        = FALSE,
  min.pct         = 0.01,
  logfc.threshold = 0.05,
  test.use        = "wilcox"
)

markers_reliable <- markers_decontX |>
  dplyr::filter(
    p_val_adj < 0.05,
    abs(avg_log2FC) >= 0.25,
    abs(pct.1 - pct.2) >= 0.10
  )

#write tables
fwrite(markers_decontX, file = "markers_despotx_all.tsv", sep="\t")

```



```{r,  fig.width=8, fig.height=7, cache=T}

usc3$celltype_despotx <- dplyr::case_when(
  usc3$clusters_despotx == "G1"  ~ "T/NK.cells",

  usc3$clusters_despotx == "G2"  ~ "Macrophages",       # Kupffer + scar-associated merged at this resolution

  usc3$clusters_despotx == "G3"  ~ "Hep.pericentral.Zone3",   # CYP2E1, GLUL, ADH1A/4/6, lipogenesis
  usc3$clusters_despotx == "G4"  ~ "Hep.periportal.Zone1.1",  # ASS1, PCK1, SDS, HAMP, ARG1
  usc3$clusters_despotx == "G5"  ~ "Hep.midzonal.Zone2",      # weak/intermediate on both axes
  usc3$clusters_despotx == "G8"  ~ "Hep.IFNg.stimulated",     # zonation washed out by CXCL9/10, GBP1/4, STAT1
  usc3$clusters_despotx == "G12" ~ "Hep.periportal.Zone1.2",  # periportal, secondary population
  usc3$clusters_despotx == "G14" ~ "Hep.periportal.Zone1.3",  # periportal, third population (CEBPA+ metabolic)

  usc3$clusters_despotx == "G6"  ~ "Fibroblasts.1",     # activated myofibroblast
  usc3$clusters_despotx == "G7"  ~ "Fibroblasts.2",     # quiescent HSC/portal

  usc3$clusters_despotx == "G9"  ~ "Endothelials",

  usc3$clusters_despotx == "G10" ~ "Cholangiocytes",

  usc3$clusters_despotx == "G11" ~ "Proliferating",

  usc3$clusters_despotx == "G13" ~ "Plasma.cells.1",    # kappa-dominant
  usc3$clusters_despotx == "G16" ~ "Plasma.cells.2",    # lambda-dominant

  usc3$clusters_despotx == "G15" ~ "Monocytes.Neutrophils",

  TRUE ~ "Undefined"
)

celltype_colors_despotx <- c(
  "T/NK.cells"    = "#33A02C",
  "Macrophages"   = "#E6550D",

  # Hepatocytes — blue shades by zone, IFN-activated pulled out in red to flag it as non-zonal
  "Hep.pericentral.Zone3"  = "#08306B",
  "Hep.periportal.Zone1.1" = "#2171B5",
  "Hep.periportal.Zone1.2" = "#4292C6",
  "Hep.periportal.Zone1.3" = "#6BAED6",
  "Hep.midzonal.Zone2"     = "#9ECAE1",
  "Hep.IFNg.stimulated"    = "#B2182B",

  "Fibroblasts.1" = "#6A3D9A",
  "Fibroblasts.2" = "#A27AC3",

  "Endothelials"          = "#00A087",
  "Cholangiocytes"        = "#D8A600",
  "Proliferating"         = "#525252",
  "Plasma.cells.1"        = "#C51B7D",
  "Plasma.cells.2"        = "#DE77AE",
  "Monocytes.Neutrophils" = "#CB181D",

  "Undefined" = "#BDBDBD"
)


DimPlot(
  usc3, 
  group.by = "clusters_despotx",
  label = T,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages scPearson Batch correction -> despotX applied") +
  theme(legend.position = "none")

DimPlot(
  usc3, 
  group.by = "celltype_despotx",
  label = T,
  raster = T,
  label.size = 2.5,
  repel = T,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP macrophages scPearson Batch correction -> deSpotX applied") +
  theme(legend.position = "none")

```

Let's subset the T/NK cells

```{r tcells_selection, fig.width=8, fig.height=7, cache=T}

usc3_tcells = subset(
  usc3,
  (clusters %in% c("G1"))
)

DefaultAssay(usc3_tcells) = "RNA"

DimPlot(
  usc3_tcells, 
  group.by = "celltype_despotx",
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```



```{r tcells_scPearson_dimReduc, fig.width=8, fig.height=7, cache=T}

tc <- Matrix::colSums(usc3_tcells[["RNA"]]$counts)
genefreq_batch <- 
gene_frequency(usc3_tcells[["RNA"]]$counts, obs = data.table(usc3_tcells@meta.data)[,.(cell_id, sample_id)]
               ,cellid_colname = "cell_id"
               ,batch_variable = "sample_id")

usc3_tcells = FindVariableFeatures(usc3_tcells, nfeatures = 6125)
hvgs <- VariableFeatures(usc3_tcells, assay = "RNA")

### Returns a Seurat-style DimReduc object with 
pcaobj_batch <- 
sparse_quasipoisson_pca_seurat_batch(usc3_tcells[["RNA"]]$counts[hvgs,]
                                     ,totalcounts = tc
                                     ,grate = genefreq_batch[hvgs,]
                                     ,obs =data.table(usc3_tcells@meta.data)[,.(cell_id, sample_id)] 
                                     ,batch_variable = "sample_id"
                                     ,cellid_colname = "cell_id"
                                     ,scale.max = 10
                                     ,do.scale = TRUE
                                     ,do.center = TRUE,
                                     ncores = 8)

#perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc3_tcells[["pearsonpca_despotx"]] <- pcaobj_batch$reduction.data
usc3_tcells[["pearsonumap_despotx"]] <- umapobj$ump  
usc3_tcells[["pearsongraph_despotx"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc3_tcells <- Seurat::FindClusters(usc3_tcells, resolution = 1.5, graph = "pearsongraph_despotx")
usc3_tcells@meta.data$clusters_tcells <- usc3_tcells@meta.data$pearsongraph_despotx_res.1.5

DimPlot(
  usc3_tcells, 
  group.by = "sample_id",
  label = F,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells  — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

DimPlot(
  usc3_tcells, 
  group.by = "clusters_tcells",
  label = T,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_despotx") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells  — RNA, scPearson Batch correction") +
  theme(legend.position = "none")

```

Let's perform now another round of decontamintaion of this subset

```{r immune_despotX,  fig.width=4, fig.height=4, cache=T}

counts_rna <- GetAssayData(usc3_tcells, assay = "RNA", layer = "counts")   # genes x cells
dcx <- celda::decontX(
  x    = as(counts_rna, "CsparseMatrix"),
  z    = as.character(usc3_tcells$clusters_tcells),                                # cluster anchors
  seed = 1
)

usc3_tcells$decontX_contamination <- dcx$contamination
usc3_tcells[["decontX"]] <- CreateAssay5Object(counts = dcx$decontXcounts)

#delete tmp_variables....
rm(counts_rna)
rm(dcx)

# contamination on the Harmony UMAP + per-cluster distribution
FeaturePlot(
  usc3_tcells, "decontX_contamination", 
  reduction = "pearsonumap_despotx",
  assay = "decontX") +
  ggtitle("DecontX contamination")

ggplot(
  usc3_tcells@meta.data, aes(clusters_tcells, decontX_contamination, fill = clusters_tcells)) +
  geom_boxplot(outlier.size = 0.1) + theme_clean(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(x = NULL, y = "contamination", title = "Contamination by cluster")


```


```{r tcells_dimreduc_decontX, fig.width=4, fig.height=4, cache=T}

DefaultAssay(usc3_tcells) = "decontX"

tc <- Matrix::colSums(usc3_tcells[["decontX"]]$counts)
genefreq_batch <- 
  gene_frequency(usc3_tcells[["decontX"]]$counts, obs = data.table(usc3_tcells@meta.data)[,.(cell_id, sample_id)]
                 ,cellid_colname = "cell_id"
                 ,batch_variable = "sample_id")

usc3_tcells = FindVariableFeatures(usc3_tcells, nfeatures = 6215)
hvgs <- VariableFeatures(usc3_tcells, assay = "RNA")

### Returns a Seurat-style DimReduc object with 
pcaobj_batch <- 
  sparse_quasipoisson_pca_seurat_batch(usc3_tcells[["decontX"]]$counts[hvgs,]
                                       ,totalcounts = tc
                                       ,grate = genefreq_batch[hvgs,]
                                       ,obs =data.table(usc3_tcells@meta.data)[,.(cell_id, sample_id)] 
                                       ,batch_variable = "sample_id"
                                       ,cellid_colname = "cell_id"
                                       ,scale.max = 10
                                       ,do.scale = TRUE
                                       ,do.center = TRUE,
                                       ncores = 8)
#perform UMAP...
umapobj <- scPearsonPCA::make_umap(pcaobj_batch)

#Integration...
usc3_tcells[["pearsonpca_decontX"]] <- pcaobj_batch$reduction.data
usc3_tcells[["pearsonumap_decontX"]] <- umapobj$ump  
usc3_tcells[["pearsongraph_decontX"]] <- Seurat::as.Graph(umapobj$grph) 

# Clustering
usc3_tcells <- Seurat::FindClusters(usc3_tcells, resolution = 2, graph = "pearsongraph_decontX", algorithm = 4)
usc3_tcells@meta.data$clusters_tcells_decontx <- paste0(
  "G",
  usc3_tcells@meta.data$pearsongraph_decontX_res.2)

DimPlot(
  usc3_tcells, 
  group.by = "sample_id",
  label = F,
  label.size = 3,
  raster = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

DimPlot(
  usc3_tcells, 
  group.by = "clusters_tcells_decontx",
  label = T,
  raster = T,
  label.size = 3,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP Tcells scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```


```{r tcells_markers,  fig.width=4, fig.height=9, cache=T}

# Set default reductions
Idents(usc3_tcells) = usc3_tcells$clusters_tcells_decontx

usc3_tcells <- NormalizeData(
  usc3_tcells,
  assay = "RNA",
  normalization.method = "RC",
)

markers_decontX <- FindAllMarkers(
  usc3_tcells,
  assay           = "RNA",
  only.pos        = FALSE,
  min.pct         = 0.01,
  logfc.threshold = 0.05,
  test.use        = "wilcox"
)

markers_reliable <- markers_decontX |>
  dplyr::filter(
    p_val_adj < 0.05,
    abs(avg_log2FC) >= 0.25,
    abs(pct.1 - pct.2) >= 0.10
  )

#write tables
fwrite(markers_decontX, file = "markers_decontX_tcells_despotX.tsv", sep="\t")


# top_markers = markers_reliable %>%
#   group_by(cluster) %>%
#   slice_head(n=30) %>%
#   data.table()
# 
# # show Heatmap of markers
# markers.mat = Seurat::AverageExpression(
#   usc3_tcells,
#   features = top_markers$gene %>% unique(),
#   assay = "RNA", slot = "counts"
# )$RNA %>% data.frame()
# 
# pheatmap::pheatmap(markers.mat, scale="row", fontsize_row = 5)

```



```{r hieratype_tcells,  fig.width=10, fig.height=5, cache=T}

data("markerslist_tcellmajor")
data(pipeline_tcell)

fct <- clusterwise_foldchange_metrics(counts = usc3_tcells[["decontX"]]$counts[hvgs,],
                                      ,metadata = usc3_tcells@meta.data,
                                      ,cluster_column = "clusters_tcells_decontx")
idxs <- unique(unname(c(unlist(lapply(HieraType::markerslist_cd4tminor, "[[", "index_marker"))
               ,unlist(lapply(HieraType::markerslist_cd8tminor, "[[", "index_marker")))))

nms <- c("cd4t", names(HieraType::markerslist_cd4tminor), "cd8t", names(HieraType::markerslist_cd8tminor))
hmsubtype <- marker_heatmap(
  fct, 
  featsuse = unique(c("CD8A", "CD8B", "CD4", idxs)), 
  fold_change_min = 0.05, 
  cluster_prop_min = 0.05, 
  topleft_to_bottomright = F) ;print(hmsubtype)


```




```{r tcells_annotation, fig.width=8, fig.height=7, cache=T}

usc3_tcells$celltype_tcell <- dplyr::case_when(
  # confident, multi-gene-coherent calls
  usc3_tcells$clusters_tcells_decontx == "G3"  ~ "cd8_tem_cytotoxic",   # strongest — all genes in both cd8_tem & cd8_cytotoxic coherent
  usc3_tcells$clusters_tcells_decontx == "G8"  ~ "cd8_cytotoxic_ifng",  # cd8_cytotoxic coherent; also drives the CXCL9/10/STAT1 signal itself
  usc3_tcells$clusters_tcells_decontx == "G7"  ~ "cd8_exhausted.1",      # PDCD1/LAG3/TIGIT/HAVCR2/TOX all coherent
  usc3_tcells$clusters_tcells_decontx == "G4"  ~ "cd4_cd8_tcm.1",      # CCR7/IL7R/SELL coherent — tcm shared signature, lineage not separable
  usc3_tcells$clusters_tcells_decontx == "G6"  ~ "cd4_cd8_tcm.2",      # same tcm signature, stronger than G4

  # tentative — coherent multi-gene signal, but small clusters (underpowered for significance testing)
  usc3_tcells$clusters_tcells_decontx == "G13" ~ "cd4_tem",       # n=760, CD44/CCR6/CXCR3/IL7R coherent
  usc3_tcells$clusters_tcells_decontx == "G2"  ~ "cd4_th2",       # tiny cluster, GATA3/IL4/IL5/CCR4 coherent
  usc3_tcells$clusters_tcells_decontx == "G14" ~ "cd8_exhausted.2", # n=172, PDCD1/LAG3/HAVCR2/TOX coherent

  # contamination
  usc3_tcells$clusters_tcells_decontx == "G1"  ~ "Hep.Contaminant.1",  # hepatocyte
  usc3_tcells$clusters_tcells_decontx == "G9"  ~ "Hep.Contaminant.2",  # hepatocyte + dissociation-stress overlay
  usc3_tcells$clusters_tcells_decontx == "G12" ~ "Hep.Contaminant.3",  # hepatocyte + dissociation-stress overlay
  usc3_tcells$clusters_tcells_decontx == "G10" ~ "Fib.Contaminant.4",  # fibroblast
  usc3_tcells$clusters_tcells_decontx == "G11" ~ "Plasm.Contaminant.5",  # plasma cells (dominant Ig signal; may carry minor T-cell doublet signal)

  # unresolvable — dominated by technical noise despite containing confirmed real T cells (TRAV19/TRAV29DV5)
  usc3_tcells$clusters_tcells_decontx == "G5"  ~ "Undefined",

  TRUE ~ "Undefined"
)

celltype_colors_tcell <- c(
  # CD8 lineage — blue shades (darker = higher confidence)
  "cd8_tem_cytotoxic"       = "#08306B",  # G3, strongest
  "cd8_cytotoxic_ifng"      = "#08519C",  # G8
  "cd8_exhausted.1"           = "#3182BD",  # G7
  "cd8_exhausted.2" = "#9ECAE1",  # G14, small n

  # CD4 lineage — pink/purple shades
  "cd4_tem" = "#C994C7",  # G13, small n
  "cd4_th2"  = "#DE77AE",  # G2, tiny n

  # shared/mixed tcm — green shades
  "cd4_cd8_tcm.1" = "#41AB5D",  # G4
  "cd4_cd8_tcm.2" = "#1B7837",  # G6, stronger signal

  # contamination — grayscale
  "Hep.Contaminant.1" = "#D9D9D9",
  "Hep.Contaminant.2" = "#BDBDBD",
  "Hep.Contaminant.3" = "#A3A3A3",
  "Fib.Contaminant.4" = "#898989",
  "Plasm.Contaminant.5" = "#6E6E6E",

  "Undefined" = "#000000"
)


DimPlot(
  usc3_tcells, 
  group.by = "clusters_tcells_decontx",
  # cols = celltype_colors,
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

DimPlot(
  usc3_tcells, 
  group.by = "celltype_tcell",
  # cols = celltype_colors,
  label = T,
  raster = T,
  label.size = 3,
  repel = T,
  reduction = "pearsonumap_decontX") +
  theme_clean(base_size = 8) +
  ggtitle("UMAP — scPearson Batch correction -> decontX applied") +
  theme(legend.position = "none")

```



```{r hieratype_ctypes_tcells, fig.width=10, fig.height=5, cache=T}


data("markerslist_tcellmajor")
data(pipeline_tcell)

fct <- clusterwise_foldchange_metrics(counts = usc3_tcells[["decontX"]]$counts[hvgs,],
                                      ,metadata = usc3_tcells@meta.data,
                                      ,cluster_column = "celltype_tcell")
idxs <- unique(unname(c(unlist(lapply(HieraType::markerslist_cd4tminor, "[[", "index_marker"))
               ,unlist(lapply(HieraType::markerslist_cd8tminor, "[[", "index_marker")))))

nms <- c("cd4t", names(HieraType::markerslist_cd4tminor), "cd8t", names(HieraType::markerslist_cd8tminor))
hmsubtype <- marker_heatmap(
  fct, 
  featsuse = unique(c("CD8A", "CD8B", "CD4", idxs)), 
  fold_change_min = 0.05, 
  cluster_prop_min = 0.05, 
  topleft_to_bottomright = F) ;print(hmsubtype)



```








