---
title: "CosMx Spatial Transcriptomics — TA649"
author: "F. Ake"
date: "2026-05-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 7,
  engine.path = list(bash = "/usr/bin/bash")
)
```

---

## Ideas & Tasks to Develop

> *Working notes from team meetings — not part of the analysis pipeline.*

### Cell-Type Annotation

- **Transfer learning from Nanostring liver reference** — use the publicly available Nanostring liver reference dataset to perform label transfer onto our data (e.g. via `Seurat::FindTransferAnchors` / `TransferData`, or `Azimuth`). This would provide an objective, reference-anchored cell-type call as a complement or sanity-check against unsupervised clustering.

- **Manual annotation + CellMarker 2.0** — refine cluster labels manually by cross-referencing marker genes against [CellMarker 2.0](http://cellmarker.bioinf.seu.edu.cn/) and liver-specific literature. Key AIH-relevant populations to resolve: pericentral vs periportal hepatocytes, Kupffer cells, HSCs, plasma cell subtypes, cholangiocytes, and IFN-stressed hepatocytes.

### Spatial Visualisation

- **Cell segmentation plot function** — develop a `plotSegmentation(obj, col.by, ...)` function that reads polygon vertex coordinates stored in `obj@misc$polygons` (or directly from Seurat FOV spatial data) and renders filled cell outlines with `ggplot2` (`geom_polygon`), colouring each cell by any `@meta.data` column. Useful for spatially contextualising cluster assignments and QC flags. *Discuss feasibility and design with Karan before implementing.*

---

## 1. Libraries & Configuration

Load all required packages, define analysis constants (data paths, QC thresholds), and configure parallel execution.

```{r libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(future)
library(future.apply)
library(data.table)
library(openxlsx)
library(clustree)
library(ggthemes)
library(Polychrome)
library(scales)
library(colorspace)
library(scPearsonPCA)
library(ggrepel)

source("src/custom_functions.R")

# Paths & panel
COSMX_INPUTFILES_REPO        <- "/scratch/users/franzake/VIVEK_LAB/data/flatFiles"
PANEL_FOV                    <- "Hs_6k"
FOV_MAX_PROP_LOSS            <- 0.6

# QC thresholds
COUNTS_THRESHOLD_MIN          <- 30
COUNTS_THRESHOLD_MAX          <- NULL
AREA_THRESHOLD_MIN            <- NULL
AREA_THRESHOLD_MAX            <- 80e3
SBR_THRESHOLD_MIN             <- 0
SBR_THRESHOLD_MAX             <- 7.5
FOV_INTEGRITY_PERCENT_THRESHOLD <- 0.6
SPLIT_RATIO_THRESHOLD_MIN     <- NULL
SPLIT_RATIO_THRESHOLD_MAX     <- log2(4)

# Parallelise heavy steps (SCTransform, RCTD, etc.)
plan("multicore", workers = 6)
options(future.globals.maxSize = 60 * 1024^3)
```

---

## 2. Input Files

Preview the flat-file directory tree to confirm all expected CosMx output files are present.

```{r list_files}
system("/oak/stanford/groups/bhowitt/conda_envs/workspace/bin/eza -T -L 4 .")
```

---

## 3. Data Loading

Read raw CosMx flat files into a Seurat object via `readCosMx()`. Tissue condensation runs automatically, and the TMA layout is visualised by FOV.

```{r data_loading}
usc <- readCosMx(experiment_repo = COSMX_INPUTFILES_REPO)
usc

TmaPlot(usc, col.by = "FOV")
```

---

## 4. Quality Control

QC proceeds in three stages: (1) compute per-cell metrics, (2) inspect distributions and apply thresholds, and (3) map each FOV to its TMA core.

### 4.1 QC Metrics

Four metrics are used to filter low-quality cells:

- **nCount vs nFeature RNA** — transcriptomic complexity; very low or high values flag empty segmentations, background, or merged cells.
- **nCount vs Area** — segmentation consistency; unusually small or large cells reflect debris or merged-cell artefacts.
- **SBR vs nCount** — signal-to-background ratio; low values indicate noisy or unreliable transcript detections.
- **Split Ratio vs Area** — boundary fragmentation; high values indicate over-segmentation and poor transcript assignment.

```{r qc_controls_and_filtering}
# Compute per-cell QC metrics
usc <- computeFOVintegrity(usc, max_prop_loss = FOV_MAX_PROP_LOSS, panel_name = PANEL_FOV)
usc <- computeSBR(usc)
usc <- computeSplitRatio(usc)
usc$SplitRatioToLocal_log2 <- log2(usc$SplitRatioToLocal)

# Optional spatial overlays (uncomment to inspect)
# TmaPlot(usc, col.by = "fov_signal_loss", pt.size = 0.01)
# TmaPlot(usc, col.by = "flag_fov_integrity")
# TmaPlot(usc, col.by = "log2SBR")
# TmaPlot(usc, col.by = "SplitRatioToLocal_log2")

# Raw QC scatter plots with thresholds
plots <- plotQCs(
  usc,
  count_min       = COUNTS_THRESHOLD_MIN,
  count_max       = COUNTS_THRESHOLD_MAX,
  area_min        = AREA_THRESHOLD_MIN,
  area_max        = AREA_THRESHOLD_MAX,
  sbr_min         = SBR_THRESHOLD_MIN,
  sbr_max         = SBR_THRESHOLD_MAX,
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX,
  base_size = 15) %>%
  patchwork::wrap_plots(nrow = 4)
plots

# Apply QC filtering
usc_qc <- QCfiltering(
  usc,
  count_min               = COUNTS_THRESHOLD_MIN,
  count_max               = COUNTS_THRESHOLD_MAX,
  area_min                = AREA_THRESHOLD_MIN,
  area_max                = AREA_THRESHOLD_MAX,
  fov_integrity_threshold = FOV_INTEGRITY_PERCENT_THRESHOLD,
  sbr_min                 = SBR_THRESHOLD_MIN,
  sbr_max                 = SBR_THRESHOLD_MAX,
  split_ratio_min         = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max         = SPLIT_RATIO_THRESHOLD_MAX)

# Post-filter QC scatter plots
plots_qcs <- plotQCs(
  usc_qc,
  count_min       = COUNTS_THRESHOLD_MIN,
  count_max       = COUNTS_THRESHOLD_MAX,
  area_min        = AREA_THRESHOLD_MIN,
  area_max        = AREA_THRESHOLD_MAX,
  sbr_min         = SBR_THRESHOLD_MIN,
  sbr_max         = SBR_THRESHOLD_MAX,
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX,
  base_size = 15) %>%
  patchwork::wrap_plots(nrow = 4)
plots_qcs

# UpSet plot — overlap of filtering flags
usc_qc@misc$qc_plots$upset
```

### 4.2 FOV-to-Core Assignment

Parse the TMA construction-key Excel file, map each FOV to its core grid position, then propagate core identities to individual cells. One FOV (182) requires a manual correction.

```{r assignation_of_fov_to_cores}
# Parse TMA map (transpose: rows = TMA rows, cols = TMA cols)
tma_mat <- parseTMAmap("data/flatFiles/TA649/TA-649 Autoimmune Hepatitis biopsy.xlsx") %>% t()
tma_mat[is.na(tma_mat)] <- "control"
tma_mat

# Assign FOVs to TMA cores
usc_qc <- assignCoresToCells(usc_qc,
  tma_map              = tma_mat,
  fov_positions        = as.data.frame(fread("data/flatFiles/TA649/TA649_fov_positions_file.csv.gz")),
  core_drift_tolerance = 0.40)

usc_qc@misc$fov_core_plots$TA649$Whole_Slide_by_Row
usc_qc@misc$fov_core_plots$TA649$Whole_Slide_by_Column

# Manual reassignment: FOV 182 was misassigned — correct to C2R3 / sample 37170
usc_qc$core_str  <- (usc_qc@meta.data %>% dplyr::mutate(core_str  = ifelse(fov == 182, "C2R3",  core_str)))$core_str
usc_qc$sample_id <- (usc_qc@meta.data %>% dplyr::mutate(sample_id = ifelse(fov == 182, "37170", sample_id)))$sample_id

# Visualise sample assignments on TMA
sample_cols <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00",
  "#984EA3", "#A65628", "#F781BF", "#FFED6F",
  "#00CED1", "#8B0000", "#006400", "#FF6347",
  "#4B0082", "#2E8B57"
)
TmaPlot(usc_qc, col.by = "sample_id", label = TRUE, cols = sample_cols, dark = TRUE)
```

---

## 5. Dimensionality Reduction

Four embedding strategies are compared to find the best representation for downstream clustering. Batch effects across samples are addressed progressively.

### 5.1 SCTransform — No Batch Correction

Baseline embedding using SCT normalisation, PCA, and UMAP. Exposes any sample-level batch structure.

```{r dimensionality_reduction_noBatch_correction}
usc_qc <- FindVariableFeatures(usc_qc, nfeatures = 2000)

usc_qc <- SCTransform(
  usc_qc,
  assay           = "RNA",
  vars.to.regress = c("nCount_RNA", "nFeature_RNA"),
  clip.range      = c(-10, 10))

usc_qc <- RunPCA(usc_qc, npcs = 30, assay = "SCT")
ElbowPlot(usc_qc, ndims = 50)
NPCS <- 10

usc_qc <- RunUMAP(usc_qc, dims = 1:NPCS, reduction = "pca")
UMAP.plot0 <- DimPlot(usc_qc, group.by = "sample_id", label = TRUE, reduction = "umap") +
  ggtitle("UMAP — SCT, no batch correction")
UMAP.plot0
```

A clear sample-level batch effect is visible in the UMAP.

### 5.2 SCTransform + Harmony Batch Correction

Harmony integration applied on the SCT PCA embedding to correct sample-level batch.

```{r DR_with_harmony_batch_correction}
tmp_00 <- usc_qc[["pca"]]
harmony_embedding1 <- harmony::RunHarmony(
  usc_qc[["pca"]]@cell.embeddings,
  meta_data = usc_qc@meta.data,
  vars_use  = "sample_id")
tmp_00@cell.embeddings  <- harmony_embedding1
usc_qc[["pca_harmony"]] <- tmp_00

usc_qc[["umap_sct_harmony"]] <- (RunUMAP(usc_qc, reduction = "pca_harmony", dims = 1:NPCS))@reductions$umap
UMAP.plot1 <- DimPlot(usc_qc, group.by = "sample_id", label = TRUE, reduction = "umap_sct_harmony", pt.size = 1.5) +
  ggtitle("UMAP — SCT + Harmony batch correction")
UMAP.plot1
```

A residual "globe" artefact in the UMAP suggests SCT normalisation is sub-optimal for this dataset. The quasi-Poisson approach below addresses this.

### 5.3 scPearsonPCA — Native Batch Integration

Quasi-Poisson PCA with built-in per-batch gene-frequency correction. More statistically appropriate for count-based spatial data.

```{r scPearson_approach_with_batch_integration}
DefaultAssay(usc_qc) <- "RNA"

tc <- Matrix::colSums(usc_qc[["RNA"]]$counts)

genefreq_batch <- gene_frequency(
  usc_qc[["RNA"]]$counts,
  obs            = data.table(usc_qc@meta.data)[, .(cell_ID, sample_id)],
  cellid_colname = "cell_ID",
  batch_variable = "sample_id")

tmp_00 <- sparse_quasipoisson_pca_seurat_batch(
  usc_qc[["RNA"]]$counts[hvgs, ],
  totalcounts    = tc,
  grate          = genefreq_batch[hvgs, ],
  obs            = data.table(usc_qc@meta.data)[, .(cell_ID, sample_id)],
  batch_variable = "sample_id",
  cellid_colname = "cell_ID",
  scale.max      = 10,
  do.scale       = TRUE,
  do.center      = TRUE)
usc_qc[["pearson_pca_batch"]] <- tmp_00$reduction.data

tmp_01 <- scPearsonPCA::make_umap(tmp_00)
usc_qc[["pearson_umap_batch"]]  <- tmp_01$ump
usc_qc[["pearson_graph_batch"]] <- Seurat::as.Graph(tmp_01$grph)

UMAP.plot2 <- DimPlot(usc_qc, reduction = "pearson_umap_batch", group.by = "sample_id", pt.size = 1.5) +
  ggtitle("UMAP — scPearsonPCA, native batch integration")
UMAP.plot2
```

### 5.4 scPearsonPCA + Harmony Batch Correction

Quasi-Poisson PCA followed by Harmony correction as an additional comparison.

```{r scPearson_approach_with_harmony_batch_on_pearson_pca}
DefaultAssay(usc_qc) <- "RNA"

tc       <- Matrix::colSums(usc_qc[["RNA"]]$counts)
genefreq <- scPearsonPCA::gene_frequency(usc_qc[["RNA"]]$counts)

usc_qc <- Seurat::FindVariableFeatures(usc_qc, nfeatures = 2000)
hvgs   <- VariableFeatures(usc_qc)

tmp_02 <- sparse_quasipoisson_pca_seurat(
  usc_qc[["RNA"]]$counts[hvgs, ],
  totalcounts = tc,
  grate       = genefreq[hvgs],
  scale.max   = 10,
  do.scale    = TRUE,
  do.center   = TRUE)

harmony_embedding2 <- harmony::RunHarmony(
  tmp_02$reduction.data@cell.embeddings,
  meta_data = usc_qc@meta.data,
  vars_use  = "sample_id")
tmp_02$reduction.data@cell.embeddings <- harmony_embedding2

tmp_03 <- scPearsonPCA::make_umap(tmp_02)
usc_qc[["pearson_pca_harmony"]] <- tmp_02$reduction.data

usc_qc[["pearson_umap_sct_harmony"]] <- (RunUMAP(usc_qc, reduction = "pearson_pca_harmony", dims = 1:13))@reductions$umap
UMAP.plot3 <- DimPlot(usc_qc, group.by = "sample_id", label = TRUE, reduction = "pearson_umap_sct_harmony", pt.size = 1.5) +
  theme(legend.position = "none") +
  ggtitle("UMAP — scPearsonPCA + Harmony batch correction")
UMAP.plot3
```

#### QC Overlay on Embedding

Inspect whether low-quality cells cluster together in the chosen embedding.

```{r more_on_QC}
lapply(c("nCount_RNA", "nFeature_RNA"), function(x) {
  FeaturePlot(usc_qc, feature = x, order = TRUE, raster = FALSE, assay = "SCT") + ggthemes::theme_clean()
}) %>% patchwork::wrap_plots(nrow = 2)
```

---

## 6. Clustering & Cell-Type Annotation

### 6.1 Marker Discovery

Find top differentially expressed markers per cluster, then inspect cluster-level average expression with a heatmap.

```{r marker_discovery}
Idents(usc_qc) <- usc_qc$seurat_clusters
usc.markers <- data.table(FindAllMarkers(usc_qc, assay = "SCT", only.pos = TRUE))
usc.markers

top5.markers <- (usc.markers %>% group_by(cluster) %>% slice_head(n = 30)) %>% data.table()

avg.mat <- data.frame(AverageExpression(
  usc_qc[top5.markers$gene, ],
  group.by = "seurat_clusters",
  layer    = "counts")[["SCT"]])
pheatmap::pheatmap(avg.mat, scale = "row", display_numbers = TRUE)
```

### 6.2 Cell-Type Annotation

Project known liver cell-type marker panels onto the UMAP to guide cluster labelling.

```{r feature_plots_cell_types}
jci_insight_markers <- list(
  "Kupffer_cells/Macrophages"     = c("C1QC", "C1QB", "C1QA", "MARCO"),
  "Hepatocytes"                   = c("SLC27A5", "ADH4", "SLC2A2", "ARG1", "CYP4A11"),
  "HSCs"                          = c("COL1A1", "COL3A1", "DCN", "TAGLN", "SPARC"),
  "Periportal_hepatocytes"        = c("SAA1", "CRP", "FBP1", "ASS1", "PCK1"),
  "Epithelial_cells"              = c("KRT18", "PTK2", "SIGMAR1", "VPS29", "MED15"),
  "Stressed_midzonal_hepatocytes" = c("HAMP", "FOS", "JUND", "SERPINE1", "CXCL2"),
  "Plasma_cells_1"                = c("IGKC", "IGHG1/2", "JCHAIN", "MZB1", "IGHA1"),
  "IFN_stressed_hepatocytes"      = c("CXCL9", "CXCL10", "STAT1", "HLA-G", "HLA-F"),
  "Plasma_cells_2"                = c("IGLC1/2", "IGLL5", "IGKC", "IGHG1/2", "JCHAIN"),
  "Pericentral_hepatocytes"       = c("GLUL", "TDO2", "FKBP5", "MGST1", "HSD11B1"),
  "Cholangiocytes"                = c("KRT7", "KRT8", "TACSTD2", "MMP7", "SPP1")
)

plots <- lapply(names(jci_insight_markers)[1:2], function(x) {
  message(x)
  p <- FeaturePlot(usc_qc, features = jci_insight_markers[[x]], order = TRUE, raster = FALSE, combine = FALSE)
  p <- lapply(p, function(x) x + theme(legend.position = "none"))
  patchwork::wrap_plots(p, nrow = 1)
})

patchwork::wrap_plots(plots, nrow = 2)
```
