---
title: "Testing CosMx analysis"
output: github_document
author: "Jose Liboy Lugo"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r Loading libraries and setting up directories}
library(Seurat)
library(dplyr)
library(ggplot2)
library(data.table)

date <- Sys.Date()
source("cosmxqc_functions.r")

                         
COSMX_INPUTFILES_REPO        <- "/scratch/users/jliboy/FT/JL01/Flatfiles"
COSMX_OUTPUTFILES_REPO       <- "/home/users/jliboy/research_projects/FT/JL01/prelim_data/"
PANEL_FOV                    <- "Hs_UCC"
PANEL_PLEX                   <- 1000
NEG_PANEL_PLEX               <- 10
FOV_MAX_PROP_LOSS            <- 0.6

# QC thresholds
COUNTS_THRESHOLD_MIN          <- 20
COUNTS_THRESHOLD_MAX          <- NULL
AREA_THRESHOLD_MIN            <- NULL
AREA_THRESHOLD_MAX            <- 60e3
SBR_THRESHOLD_MIN             <- 0
SBR_THRESHOLD_MAX             <- NULL
FOV_INTEGRITY_PERCENT_THRESHOLD <- 0.6
SPLIT_RATIO_THRESHOLD_MIN     <- NULL
SPLIT_RATIO_THRESHOLD_MAX     <- log2(4)
```

## Loading merged seurat object. 

In this section, CosMx seurat objects are loaded into R using the `readCosMxSeurat` function. 
The resulting object, `ft`, is a Seurat object that contains the data and metadata for the CosMx experiment.

```{r dataprep} 
options(future.globals.maxSize = 100 * 1024^3) 
ft <- readCosMxSeurat(experiment_repo = COSMX_INPUTFILES_REPO)
ft
```

## Plotting FOV QC metrics to evaluate data quality before running cell QC

``` {r plotting data qc per fov}

#Plotting mean transcript per cell per fov. 
fov_positions <- fov_plotting(COSMX_INPUTFILES_REPO, ft)

#Saving fov qc plots
for (i in seq_along(fov_positions)) {
  ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_fov_qc_plots_", names(fov_positions)[[i]], ".png"), plot = fov_positions[[i]], width = 10, height = 8, units = "in")
}

#Plotting mean transcript per cell per fov boxplots
meantranscript_per_cell_per_fov_boxplot <- fov_box_plot(ft)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_meantranscript_per_cell_per_fov_boxplot.png"), plot = meantranscript_per_cell_per_fov_boxplot, width = 10, height = 8, units = "in")

#Plotting mean staining intensity by slide and marker
staining_qc_plot <- staining_qc_plot(ft)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_staining_qc_plot.png"), plot = staining_qc_plot, width = 10, height = 8, units = "in")

#Creating a slide QC table summarizing key QC metrics for each slide. 
qc_table <- slide_qc_table(ft, PANEL_PLEX, NEG_PANEL_PLEX)
write.csv(qc_table, file = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_slide_qc_table.csv"), row.names = FALSE)


```

## Runing cell and FOV QC

```{r running cell qc}

ft <- computeFOVintegrity(ft, max_prop_loss = FOV_MAX_PROP_LOSS, panel_name = PANEL_FOV)
ft <- computeSBR(ft)
ft <- computeSplitRatio(ft)
ft$SplitRatioToLocal_log2 <- log2(ft$SplitRatioToLocal)

```
## Plotting the QC results 

```{r plotting qc}
library(dplyr)
# Raw QC scatter plots with thresholds
plots <- plotQCs(
  ft,
  count_min       = COUNTS_THRESHOLD_MIN,
  count_max       = COUNTS_THRESHOLD_MAX,
  area_min        = AREA_THRESHOLD_MIN,
  area_max        = AREA_THRESHOLD_MAX,
  sbr_min         = SBR_THRESHOLD_MIN,
  sbr_max         = SBR_THRESHOLD_MAX,
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX,
  base_size = 15) %>%
  patchwork::wrap_plots(nrow = 4)

ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_qc_plots.png"), plot = plots, width = 10, height = 15, units = "in")

```

## FLagging cells based on the fov and cell QCs. 

Notice that I am not removing cells, just flagging. 

```{r flagging cells based on qc}

ft <- qc_flagging(
  ft,
  count_min               = COUNTS_THRESHOLD_MIN,
  count_max               = COUNTS_THRESHOLD_MAX,
  area_min                = AREA_THRESHOLD_MIN,
  area_max                = AREA_THRESHOLD_MAX,
  fov_integrity_threshold = FOV_INTEGRITY_PERCENT_THRESHOLD,
  sbr_min                 = SBR_THRESHOLD_MIN,
  sbr_max                 = SBR_THRESHOLD_MAX,
  split_ratio_min         = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max         = SPLIT_RATIO_THRESHOLD_MAX)

```

## Saving flagged metadata for visualization in napari

I am doing this step because I want to be able to inspect the flagging on napari. The name of the metadata file should have 
the "_metadata.csv" sufix in order to be read by the cosmx-napari software. 

``` {r saving metadata}

#Saving metadata files for each slide for napari
slidenames <- unique(ft$slidename)
md <- ft@meta.data
for (i in seq_along(slidenames)) {
  md_sub <- md %>%
  filter(slidename == slidenames[i]) %>%
  select(starts_with("flag_"), cell_ID)
  
  rownames(md_sub) <- NULL
  md_sub <- md_sub %>% relocate(cell_ID) # moves cell_ID to first column position
  write.csv(
  md_sub,
  file = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_", slidenames[i], "_metadata.csv"),
  col.names = TRUE,
  row.names = FALSE,
  quote = FALSE
)
}
```


## Removing flagged cells for downstream analysis

In this section, I am subsetting the seurat object to only include cells with 'flag_overall' == FALSE

``` {r removing flagged cells}

#remove flagged cells
ft <- subset(ft, ft$flag_overall == FALSE)
ft

```


## Mapping TMA cores with a map keys

In this section, I am subsetting the the seurat object to only process cells corresponding to the TMA. 
I have to do this because not all slides are TMAs. 

``` {r tma mapping}

#Subsetting seurat object with just the TMA slide. 
ft_tma <- subset(ft, slidename == "FTOtherCATMA")


#visualize TMA 
tma_plot <- TmaPlot(ft_tma, col.by = "FOV", dark = F)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_tma_plot.png"), plot = tma_plot, width = 8, height = 6, units = "in")

```

## Assingning patient IDs to each core

In this section, I am using the `assignCoresToCells()` function to assing patient IDs 
to each individual core. 

At the end, I removed FOVs with mislabeled cells. I figured that it is easier to remove those 
FOVs, rather than relabeling cells. 

``` {r core assignment}

tma_map <- readxl::read_excel(path = paste0(COSMX_INPUTFILES_REPO, "/FT-not brca and other ca tma maps.xlsx"), 
sheet = "FT- OTHER CA TMA Map", col_names = FALSE) 

tma_map <- as.matrix(tma_map)
tma_map <- matrix(sub(",.*", "", tma_map), nrow = nrow(tma_map), ncol = ncol(tma_map))
tma_map[tma_map == "" | tma_map == "NA" | tma_map == "NULL"] <- NA

tma_map <- tma_map[nrow(tma_map):1, ] # flip rows to match TmaPlot orientation

fov_positions <- read.csv(file = paste0(COSMX_INPUTFILES_REPO, "/FTOtherCATMA/FTOtherCATMA_fov_positions_file.csv.gz"))

ft_tma <- assignCoresToCells(obj = ft_tma, tma_map = tma_map, fov_positions = fov_positions)

fov_columns_coreplot <- ft_tma@misc$fov_core_plots$FTOtherCATMA$Whole_Slide_by_Column
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_tma_fov_column.png"), plot = fov_columns_coreplot, width = 8, height = 6, units = "in")

tma_plot <- TmaPlot(ft_tma, col.by = "sample_id_col", dark = F)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_tma_sampleid.png"), plot = tma_plot, width = 8, height = 6, units = "in")

#Filter TMA to remove mislabeled FOVs. 
exclude_fov <- c(28, 29, 30, 67, 69, 70, 71, 77, 95, 108, 109, 110, 111, 113, 114, 
                  241, 242, 243, 245, 246, 247, 248, 278, 279, 280, 281, 282, 283, 
                  284, 285, 286, 331, 332, 333, 334, 335, 336, 337, 338, 339, 340)

ft_tma <- subset(ft_tma, !(FOV %in% exclude_fov))


#Add patient ID to the original suerat object
ft$patient_id <- ft$slidename
ft$patient_id <- ft_tma$sample_id_col[match(ft_tma$global_cell_ID, ft$global_cell_ID)] #Adding sample IDs to the main Seurat object.

#Removing cells labeled as FTOtherCATMA, because these correspond to excluded FOVs. 
ft <- subset(ft, patient_id != 'FTOtherCATMA')


```

## Processing merged seurat object with log-normalization for individual slides

I am normalizing each individual slide afte splitting the seurat obejct layers. 

``` {r processing seurat object}

ft[["RNA"]] <- split(ft[["RNA"]], f = ft$slidename) #Split layers by slide name to process each slide individually.

ft <- NormalizeData(ft, normalization.method = "LogNormalize")
ft <- FindVariableFeatures(ft, selection.method = "vst", nfeatures = 900)

all.genes <- rownames(ft)
ft <- ScaleData(ft, features = all.genes)

ft <- RunPCA(ft, reduction.name = 'log_norm_pca', npcs = 50)
elbow_plot <- ElbowPlot(ft, reduction = 'log_norm_pca', ndims = 50)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_elbow_plot_lognorm.png"), plot = elbow_plot, width = 8, height = 6, units = "in")


ft <- RunUMAP(ft, reduction = "log_norm_pca",
                          reduction.name = "umap_log_norm_uncorrected",
                          n.neighbors = 30,
                          min.dist = 0.01,
                          spread = 0.3,
                          dims = 1:30)

umap_lognorm_uncorrected <- DimPlot(ft, reduction = "umap_log_norm_uncorrected", group.by = "patient_id") + ggplot2::theme_bw() + ggplot2::labs(title = "UMAP of log-normalized data (uncorrected)")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_log_norm_uncorrected.png"), plot = umap_lognorm_uncorrected, width = 8, height = 6, units = "in")


```


## Normalizing merged seurat objects with SCTranform method for individual slides

Running SCTransform to each layer (slide) individually. 

``` {r SCTransform}
ft <- SCTransform(ft, assay = 'RNA', new.assay.name = 'sct_split')

ft <- RunPCA(ft, assay = 'sct_split', reduction.name = 'sct_split_pca', npcs = 50)

elbow_plot <- ElbowPlot(ft, reduction = 'sct_split_pca', ndims = 50)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_elbow_plot_sct_split.png"), plot = elbow_plot, width = 8, height = 6, units = "in")

ft <- RunUMAP(ft, reduction = "sct_split_pca",
                          reduction.name = "umap_sct_split_uncorrected",
                          n.neighbors = 30,
                          min.dist = 0.01,
                          spread = 0.3,
                          dims = 1:40)

umap_sct_uncorrected <- DimPlot(ft, reduction = "umap_sct_split_uncorrected", group.by = "patient_id") + ggplot2::theme_bw() + ggplot2::labs(title = "UMAP of SCTransform-normalized data (uncorrected)")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_split_sct_uncorrected.png"), plot = umap_sct_uncorrected, width = 8, height = 6, units = "in")


```

## Splitting umap clusters by slide

I am doing this so I can see how cells are clustering by slide. 

``` {r umap plotting}

#Splitting umaps per slide to see if there is any specific cluster per slide. 
DimPlot(ft, reduction = "umap_sct_split_uncorrected", group.by = "patient_id", split.by = "slidename")



```

##Integrating SCTransformed data with Harmony on split seurat object

I am testing Harmony integration in SCTransformed data. 

``` {r data integration}
#Harmony integration

ft <- IntegrateLayers(object = ft, method = HarmonyIntegration, orig.reduction = "sct_split_pca",
    new.reduction = "sct_split_harmony", verbose = FALSE)

ft <- RunUMAP(ft, reduction = "sct_split_harmony",
                          reduction.name = "umap_sct_split_harmonized", 
                         n.neighbors = 30,
                          min.dist = 0.01,
                          spread = 0.3,
                          dims = 1:40)

umap_sct_split_harmony <- DimPlot(ft, reduction = "umap_sct_split_harmonized", group.by = "patient_id") + ggplot2::theme_bw() + ggplot2::labs(title = "UMAP of SCT-normalized data (harmony integration)")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_sct_split_harmonized.png"), plot = umap_sct_split_harmony, width = 8, height = 6, units = "in")


```


## Normalizing data with SCTransform for combined layers (slides)

Here I am applying a SCTransform normalization on all layers (slides) combined. 

``` {r normalization with SCTransform}

ft <- SCTransform(ft, assay = 'RNA', new.assay.name = 'SCT')


vst_model_sct <- ft[["SCT"]]@SCTModel.list$model1
gene_meta_sct <- vst_model_sct@feature.attributes



# Create a plot to evaluate residual variance
plot <- ggplot2::ggplot(gene_meta_sct, aes(x = log10(gmean), y = residual_variance)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::labs(
    x = "log10(Gene Mean Expression)",
    y = "Residual Variance",
    title = "SCTransform Residual Variance Plot"
  ) + ggplot2::geom_hline(yintercept = 1, linetype = "solid", color = "red") +
  ggplot2::theme_bw()

ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_sctransform_residual_variance_plot.png"), plot = plot, width = 8, height = 6, units = "in")

```

## Plotting PCA and UMAP plots for SCTransform-normalized data on combined slides.

``` {r pca and umap for sct normalization}
ft <- Seurat::RunPCA(ft, assay = 'SCT', reduction.name = 'sct_pca', npcs = 50)

elbow_plot <- ElbowPlot(ft, reduction = 'sct_pca', ndims = 50)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_elbow_plot_sct.png"), plot = elbow_plot, width = 8, height = 6, units = "in")

ft <- RunUMAP(ft, reduction = "sct_pca",
                          reduction.name = "umap_sct_uncorrected",
                          n.neighbors = 30,
                          min.dist = 0.01,
                          spread = 0.3,
                          dims = 1:40)

umap_sct_uncorrected <- DimPlot(ft, reduction = "umap_sct_uncorrected", group.by = "slidename") + ggplot2::theme_bw() + ggplot2::labs(title = "UMAP of SCTransform-normalized data (uncorrected)")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_sct_uncorrected.png"), plot = umap_sct_uncorrected, width = 8, height = 6, units = "in")
```


## Batch correction with Harmony on SCTransform-normalized data

Applying batch correction per patient ID. For this project, this exercise may not be needed. 

``` {r batch correction with harmony on sct normalization}

#run harmony on sct normalized data for batch correction
ft <- harmony::RunHarmony(ft, group.by.vars = "patient_id", reduction.use = "sct_pca", project.dim = FALSE)

ft <- Seurat::RunUMAP(ft, reduction = "harmony",
                          reduction.name = "umap_sct_harmonized_per_patient",
                          n.neighbors = 30,
                          min.dist = 0.01,
                          spread = 0.3,
                          dims = 1:40)

umap_sct_corrected <- DimPlot(ft, reduction = "umap_sct_harmonized_per_patient", group.by = "patient_id") + ggplot2::theme_bw() + ggplot2::labs(title = "UMAP of SCTransform-normalized data (corrected by patient)")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_sct_harmonized_per_patient.png"), plot = umap_sct_corrected, width = 8, height = 6, units = "in")
```


## Normalizing data with scPearsonPCA on combined seurat objects

Here, I am normalizing all layers combined with the scPearsonPCA package, which have proven 
quite effective with CosMx data. 

``` {r normalization with scPearsonPCA}
ft <- JoinLayers(ft, assay = "RNA") #Joining layers for split seurat object. 

DefaultAssay(ft) <- "RNA"
tc <- Matrix::colSums(ft[["RNA"]]$counts) ## total counts per cell (across all genes)
genefreq <- scPearsonPCA::gene_frequency(ft[["RNA"]]$counts) ## gene frequency (across all cells)
sum(genefreq)==1 # TRUE

ft <- FindVariableFeatures(ft, nfeatures = 900)
hvgs <- VariableFeatures(ft)

### Returns a Seurat-style DimReduc object with 
### hvgs x pcs feature loadings
### cells x pcs cell embeddings
### gene-length vector of the mean pearson of residuals
### gene-length vector of the standard deviation of pearson residuals
pcaobj <- 
scPearsonPCA::sparse_quasipoisson_pca_seurat(ft[["RNA"]]$counts[hvgs,]
                               ,totalcounts = tc
                               ,grate = genefreq[hvgs]
                               ,scale.max = 10 ## PCs reflect clipping pearson residuals > 10 SDs above the mean pearson residual
                               ,do.scale = TRUE ## PCs reflect as if pearson residuals for each gene were scaled to have standard deviation=1
                               ,do.center = TRUE ## PCs reflect as if pearson residuals for each gene were centered to have mean=0
                               ,npcs = 50)



```

## Calculating PCAs and UMAPs with scPearsonPCA-normalized data. 

Addding PCA and UMAP reductions to the seurat object. 

``` {r pca and umaps from scPearsonPCA}
ft[["pearsonpca_joined"]] <- pcaobj$reduction.data
ElbowPlot(ft, reduction = "pearsonpca_joined", ndims = 50)

umapobj <- scPearsonPCA::make_umap(pcaobj)
ft[["pearsonpca_joined"]] <- pcaobj$reduction.data
ft[["pearsonumap_joined"]] <- umapobj$ump

library(data.table)
library(ggplot2)
library(ggrepel)
umapplot <- DimPlot(ft, reduction = "pearsonumap_joined", group.by = "slidename") + 
  ggplot2::theme_bw() + ggplot2::labs(title = "UMAP of scPearsonPCA-normalized data (uncorrected)")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_pearsonpca_uncorrected.png"), plot = umapplot, width = 8, height = 6, units = "in")

split_umap <- DimPlot(ft, reduction = "pearsonumap_joined", group.by = "patient_id", split.by = "slidename")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_pearsonpca_uncorrected_sep.png"), plot = split_umap, width = 15, height = 5, units = "in")
```


## Clustering scPearsonPCA-normalized data

I will identify populations with unsupervised leiden clustering. For the FT project, 
I should see Stromal-Mesenchimal, Immune, and Tumor-Epithelial clusters. 

``` {r leiden clustering}

ft <- FindNeighbors(ft, reduction = "pearsonumap_joined", graph.name = "pearson_snn", k.param = 50, annoy.metric = "euclidean", dims = 1:2)

ft <- FindClusters(ft, graph.name = "pearson_snn", cluster.name = "pearson_leiden",
                                algorithm = 4, resolution = 0.1)  # Leiden = algorithm 4

umap_leiden = DimPlot(ft, reduction = "pearsonumap_joined", group.by = "pearson_leiden", label = TRUE, repel = TRUE, split.by = "slidename")

ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_pearsonumap_leiden.png"), plot = umap_leiden, width = 15, height = 5, units = "in")

```

## Finding top markers for each leiden cluster

``` {r finding top markers}

#Find top markers for each leiden cluster using the SCTransform-normalized data.
ft_markers <- FindAllMarkers(ft, assay = "RNA", slot = "data", group.by = "pearson_leiden", 
test.use = "wilcox", only.pos = TRUE,
 min.pct = 0.1, logfc.threshold = 0.5)

#Remove markers with p_val = 0
ft_markers <- subset(ft_markers, ft_markers$p_val_adj != 0)

#Get the top 10 markers for each leiden cluster based on avg_log2FC
ft_topmarkers <- ft_markers %>%
  group_by(cluster) %>%
  slice_head(n = 5) %>%
  ungroup() 

dotplot <- DotPlot(ft, features = unique(ft_topmarkers$gene), group.by = "pearson_leiden") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5)) +
  labs(title = "Top Markers for Each Leiden Cluster") + 
  RotatedAxis()

ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_dotplot_top_markers.png"), plot = dotplot, width = 15, height = 8, units = "in")

write.csv(ft_topmarkers, file = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_top_markers.csv"), row.names = FALSE)
```

## Exploring cell markers

``` {r feature plots}

stic_features <- c("KRT17", "COL1A1", "MK167", "TOP2A")
tcell_features <- c("IL32", "CXCR4", "CD3E", "CD2", "CD8A", "CD4")

feature_plot <- FeaturePlot(object = ft, 
features = c(stic_features),
assay = "RNA", 
slot = "data", 
reduction = "pearsonumap_joined", 
max.cutoff = 10)

ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_feature_plot_stic.png"), plot = feature_plot, width = 10, height = 10, units = "in")

```


## Performing batch correction on scPearsonPCA-normalized data

In this step, I am evaluating the effect of batch-correction based on patient ID. 
This step may not be appropriate for this project, given that I am trying to capture 
differences between patients. 

``` {r batch correction for scPearsonPCA}

genefreq_batch <- 
scPearsonPCA::gene_frequency(ft[["RNA"]]$counts, obs = data.table(ft@meta.data)[,.(global_cell_ID, slidename)]
               ,cellid_colname = "global_cell_ID"
               ,batch_variable = "slidename")

Matrix::colSums(genefreq_batch)

### Returns a Seurat-style DimReduc object with 
### hvgs x pcs feature loadings
### cells x pcs cell embeddings
### gene-length vector of the mean pearson of residuals
### gene-length vector of the standard deviation of pearson residuals
pcaobj_batch <- 
scPearsonPCA::sparse_quasipoisson_pca_seurat_batch(ft[["RNA"]]$counts[hvgs,]
                                     ,totalcounts = tc
                                     ,grate = genefreq_batch[hvgs,] ##
                                     ,obs =data.table(ft@meta.data)[,.(global_cell_ID, slidename)] 
                                     ,batch_variable = "slidename"
                                     ,cellid_colname = "global_cell_ID"
                                     ,scale.max = 10
                                     ,do.scale = TRUE
                                     ,do.center = TRUE
                                     )

umapobj_batch <- scPearsonPCA::make_umap(pcaobj_batch)
ft[["pearsonpca_batch"]] <- pcaobj_batch$reduction.data
ft[["pearsonumap_batch"]] <- umapobj_batch$ump

library(data.table)
library(ggplot2)
library(ggrepel)
umapplot_batch <- scPearsonPCA::plot_umap(umapreduc = "pearsonumap_batch", clustercol = 'slidename', semuse = ft)
umapplot_batch <- umapplot_batch + ggplot2::theme_bw() + ggplot2::labs(title = "UMAP of scPearsonPCA-normalized data (corrected by slide)")
ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_umap_pearsonpca_corrected.png"), plot = umapplot_batch, width = 8, height = 6, units = "in")
                              


```

## Saving seurat object for downstream analysis

``` {r saving seurat object}

saveRDS(ft, paste0("/scratch/users/jliboy/FT/JL01/prelim_data/", date,"_ft_processed.rds"))

```

